---
output: pdf_document
---

```{r}
library(tidyverse)
library(kableExtra)
library(RColorBrewer)
```

# Describing Data

Once data have been collected, they are typically described via
graphical and numeric means. The methods used to describe the data will
depend on its type (nominal, ordinal, or numeric). We also need to
distinguish whether the data corresponds to a sample or a population. In
this chapter, we focus purely on describing a set of measurements, not
making inferences. First we consider graphical and numeric descriptions
of a single variable. Then we consider pairs of variables.

## Graphical Description of a Single Variable

Depending on the type of measurement, common plots are **pie charts**,
**bar charts**, **histograms**, **box plots**, and **density plots**.

**Pie charts** can be used to describe any variable type. Continuous numeric
variables must be collapsed into "bins" or "buckets." The size of the
sectors of the pie represent the relative frequency of each category.

**Bar charts** are used to describe nominal or ordinal data. The variable
levels are arrayed on the bottom (or left side) of the plot and bars
above (or beside) the levels represent the frequency or relative
frequency of the number of observations belonging to the various
categories.

**Histograms** are used for numeric variables, where the heights of the bars
above the bins represent the frequency or relative frequency of the
various bins.

**Box plots** are used for numeric variables. They identify particular
percentiles of a distribution and are useful in detecting outlying
observations and spread in the distribution.

**Density plots** are used for numeric variables. They offer a smoother
description of the measurements than a histogram does and are simple to
obtain with modern statistical software packages.

**Example 2.1: Hand Postures of Blues Guitarists**

A study was conducted among $n=93$ blues guitarists that classified
the guitarists with respect to several categorical variables
[@cohen_1996].
Among the variables was "hand posture," with levels: Extended, Stacked, and Lutiform. Figure \@ref(fig:blueshand1) is
a pie chart representing the proportions of guitarists classified in the
three hand posture categories. Figure \@ref(fig:blueshand2) displays a bar chart of the same data.

```{r blueshand1, fig.cap='Pie chart of Hand Postures for $n$=93 blues guitarists', out.width='80%', fig.asp=.75, fig.align='center', echo=FALSE}

bh <- read.csv("blues_hand.csv")
head(bh, 2)
bh$handPost.f <- factor(bh$handPost, levels=1:3,
                        labels=c("Extended", "Stacked", "Lutiform"))
table.handPost <- table(bh$handPost)
names(table.handPost) <- c("Extended", "Stacked", "Lutiform")

hp.labels <- c("Extended", "Stacked", "Lutiform")

pie(table.handPost, col=c("pink", "yellow", "green"),
#    labels=as.character(hp.labels),
    main="Blues Hand Postures")

```

```{r blueshand2, fig.cap='Bar chart of Hand Postures for $n$=93 blues guitarists', out.width='80%', fig.asp=.75, fig.align='center', echo=FALSE}

## Continuation of blueshand1

barplot(table.handPost, col=c("pink", "yellow", "green"),
    main="Blues Hand Postures")
```

We see that over half of the guitarists utilized the extended hand posture,
approximately a third, used the stacked, and the remaining approximately a ninth used the lutiform.

$$\nabla$$

**Example 2.2: Measurements of the Velocity of Light circa 1931-1933**

A.A. Michelson, F.G. Pease, and F. Pearson set up an approximately one
mile long tube to make determinations of the speed of light near Irvine,
CA in the early 1930s [@michelson_1935]. Without getting into the very detailed description
given in the paper, we have 1010 determinations of the velocity of light
after having removed some runs with anomalous values in the table.
Further, we do not include weights that varied due to the experimental
protocol as it evolved during the data collection process. Figure
\@ref(fig:michelson1) provides a histogram of the $n=1010$ measurements
(approximated from their tabular information) as well as a smooth
density function overlay on the graph. The values on the graph represent
velocity - 299000 km/sec. The individual measurements are mound shaped
around a center point with arithmetic mean of 299773.5 km/sec. Modern
assessments of the velocity of light in a vacuum is 299792.5 km/sec.

```{r michelson1, fig.cap='Michelson Speed of Light Measurements (Deviations from 299000 km/sec)', out.width='80%', fig.asp=.75, fig.align='center', echo=FALSE}

mich1931 <- read.csv("michelson1931.csv")
head(mich1931,2)

hist(mich1931$velocity, freq=F, breaks=seq(690,870,5),
   col="orange", border="blue", xlab="velocity", 
   main="Relative Frequency Histogram of Velocity of Light")
lines(density(mich1931$velocity))

```

Measurements were made in 4 groups of series: Series 1-54
(2/16/1931-7/14/1931), Series 55-110 (3/3/1932-5/13/1932), Series
111-158 (5/13/1932-8/4/1932), and Series 159-233 (12/3/1932-2/27/1933).
Side-by-side box plots are given in Figure \@ref(fig:michelson2). The
box plot identifies from bottom to top the following elements.

-   Minimum: Bottom of line at bottom of plot (or the lowest circle)
-   Range for lowest 25% of measurements: Distance from minimum
    observation to the bottom of the box
-   25th percentile (Lower Quartile, aka LQ): Bottom line of box
-   Range for the 25th to 50th percent of participants: Distance between
    bottom of box and second horizontal line
-   Median (50th percentile): Second horizontal line
-   Range for the 50th to 75th percent of participants: Distance between
    second horizontal line and top of box
-   Interquartile Range (IQR): Distance between top (75th percentile)
    and bottom (25th percentile) of the box
-   75th percentile (Upper Quartile, UQ): Top line of the box
-   Range for 75th to 100th percent of participants: Distance from the
    top of the box to the maximum observation
-   Maximum: Top of line at the top of plot (or the highest circle)
-   Lower line extends either to the minimum or 1.5(IQR) below the LQ,
    whichever is shortest.
-   Upper line extends either to the maximum or 1.5(IQR) above the UQ,
    whichever is shortest.
-   Circles represent outlying measurements (very extreme measurements).

The precision of the measurements tend to improve slightly over the
course of the study. Note that the average weights for the individual
measurements included in this analysis were approximately 1.65 for the
first series and approximately 3 for the remaining series.

```{r michelson2, fig.cap='Side-by-side boxplots for velocities measured in the 4 groups of series', out.width='80%', fig.asp=.75, fig.align='center', echo=FALSE}

mich1931$SeriesGrp <- factor(mich1931$SeriesGrp)   # Make variable a factor

plot(mich1931$velocity ~ mich1931$SeriesGrp, 
     col=c("coral","green", "pink", "yellow"),
     xlab="Series Group", ylab="velocity", 
     main="Side-by-Side Boxplots of Velocity by Series Group")

```

$$ \nabla $$

**Example 2.3: Body Mass Index for National Hockey League Players - 2014/2015 Season**

Body mass index (BMI) is a measure of body fat that is based on the the
work of Adolphe Quetelet, a renowned Belgian researcher in astronomy and
statistics and other areas, particularly social sciences. The formulas
for BMI in the metric and American systems are given below.

$$ BMI = \mbox{mass(kg)/height(m)}^2 = \mbox{703$\times$mass(lbs)/height(in)}^2 $$

Data for all National Hockey League (NHL) players are obtained, reported
in pounds (lbs) and inches, respectively.
A histogram is given in Figure \@ref(fig:nhlbmi1). The histogram is
approximately symmetric and mound-shaped, centered above 26.

$$ \nabla $$

```{r nhlbmi1, fig.cap='Body Mass Index for NHL Players 2014/2015 Season', out.width='80%', fig.asp=.75, fig.align='center', echo=FALSE}

nhl <- read.csv("nhl_ht_wt_1415.csv")
#head(nhl,2)

nhl$BMI <- 703*nhl$Weight/(nhl$Height^2)

hist(nhl$BMI, breaks=30, xlab="Body Mass Index", col="green",
  main="NHL BMI Distribution 2014/2015")

```

**Example 2.4: Female and Male Speeds at Washington, DC Rock and Roll Marathon - 2015**

The 2015 Rock and Roll Marathon in Washington, D.C. was completed by
1045 female and 1454 male participants. Each participant's time to
complete the marathon was converted to a speed (miles per hour).
Histograms and kernel density plots for females and males are given in
Figure \@ref(fig:rrmarathon1), and side-by-side box plots are
given in Figure \@ref(fig:rrmarathon2). For both genders, there
tend to be more cases at lower speeds with a few extreme cases with
higher speeds. These distributions are **right-skewed**.

A smooth version of a boxplot, which does not separate the measurements
into quantiles is a **violin plot**. For the marathon data, one is
displayed in Figure \@ref(fig:rrmarathon3).



```{r rrmarathon1, fig.cap='Histograms and Densities of Velocities at Rock and Roll Marathon 2015 (mph)', out.width='80%', fig.asp=.75, fig.align='center', echo=FALSE}

rr.mar <- read.csv("rocknroll_marathon_mf2015a.csv")
head(rr.mar,2)

d.F <- density(rr.mar$mph[rr.mar$Gender=="F"])
d.M <- density(rr.mar$mph[rr.mar$Gender=="M"])

## Set up 2x2 grid for plots
par(mfrow=c(2,2))
## Histograms for Female and Male mph
hist(rr.mar$mph[rr.mar$Gender=="F"], col="green", xlab="Female Speeds",
  main="Histogram of Female Speeds")
hist(rr.mar$mph[rr.mar$Gender=="M"], col="purple", xlab="Male Speeds",
  main="Histogram of Male Speeds")
## Density Plots for Female and Male mph
plot(d.F, col="green", lwd=2, main="Kernel Density Plot for Female Speeds",    xlab="Female Speeds", ylab="Density") 
plot(d.M, col="purple", lwd=2, main="Kernel Density Plot for Male Speeds",    xlab="Male Speeds", ylab="Density")
par(mfrow=c(1,1))    ## Re-set plot grid to 1 plot
```

```{r rrmarathon2, fig.cap='Box Plots of Velocities at Rock and Roll Marathon 2015 (mph)', out.width='80%', fig.asp=.75, fig.align='center', echo=FALSE}
color1 <- c("green", "purple")
plot(factor(rr.mar$Gender), rr.mar$mph, col=color1, xlab="Gender",
     ylab="mph",
     main="Side-by-Side Boxplots mph x Gender")

```

```{r rrmarathon3, fig.cap='Violin Plots of Velocities at Rock and Roll Marathon 2015 (mph)', out.width='80%', fig.asp=.75, fig.align='center', echo=FALSE}

p <- ggplot(rr.mar, aes(y=mph, x=Gender, fill=Gender)) +   geom_violin()
p + scale_fill_manual(values=c("green", "purple"))

```

$$ \nabla $$
**Time series** plots are widely used in many areas including economics,
finance, climatology, and biology. These graphs include one or more
characteristics being observed in a sequential time order. These plots
can be based on virtually any level of sampling interval.

**Example 2.5: Miami Monthly and Annual Mean Temperature 1/1949-12/2014**

They can be used to detect trend and cyclical patterns over time. Figure \@ref(fig:miami1) shows the the monthly and annual mean
temperature in Miami for the years 1949 through 2014. Clearly there is a
cyclical pattern occurring within years, and after a flat early annual
series, there certainly appears to be evidence of an increasing trend
over approximately the second half of the series (after about 1970).

```{r miami1, fig.cap='Miami Monthly and Annual Temperatures 1949-2014', out.width='80%', fig.asp=.75, fig.align='center', echo=FALSE}

mw1 <- read.csv("miami_weather.csv")
head(mw1,2)

yearMeanTemp <- aggregate(meanTemp ~ Year, mw1, mean)

par(mfrow=c(2,1))
plot(mw1$meanTemp, type="l", col="blue", xlab="Month", ylab="Mean Temp",
     main="Miami Monthly Mean Temp (F) 1949-2014")
plot(yearMeanTemp[,1], yearMeanTemp[,2], type="l", col="blue",
     xlab="Year", ylab="Mean Temp",
     main="Miami Yearly Mean Temp (F) 1949-2014"
     )
par(mfrow=c(1,1))
```

$$ \nabla  $$

## Numerical Descriptive Measures of a Single Variable

Numerical descriptive measures describe a set of measurements in
quantitative terms. When describing a **population** of measurements,
they are referred to as **parameters**; when describing a **sample** of
data, they are referred to as **statistics**.

In terms of nominal and ordinal data, **proportions** are generally the
numeric measures of interest. These are simply the fraction of
measurements falling into the various possible levels (and must sum to
1). For ordinal variables, the **cumulative proportions** are also of
interest, representing the fraction of measurements falling in or below
the various categories.

### Measures of Central Tendency

There are two commonly reported measures of central tendency, or
location for a set of measurements. The **mean** is the sum of all
measurements divided by the number of measurements, and is reported
often as "per capita" in economic reports. The mean is the"balance
point" of a set of measurements in a physical sense. The **median** is
the point where half of the measurements fall at or below it, and half
of the measurements fall at or above it. It is also the 50th percentile
of the set of measurements. Many economic reports state median values. A
third, less reported measure is the **mode** which really is only
appropriate for discrete variables, and is the value that occurs most
often. For a histogram of discretely measured data, the mode is the
level with the highest bar.

Note that the mean is affected by outlying measurements, as it is the
sum of all measurements, evenly distributed among all of the
measurements. The median is more "robust" as it is not effected by the
actual values of individual measurements, only the center of them. The
formulas for the population mean $\mu$, based on a population of $N$
items and the sample mean $\overline{y}$ for a sample of $n$ items are
given below.

$$ \mbox{Population Mean: } \mu = \frac{\sum_{i=1}^N y_i}{N} \qquad \qquad \mbox{Sample Mean: } \overline{y} = \frac{\sum_{i=1}^n y_i}{n} $$

To obtain the median, measurements are ordered from smallest to largest,
and the middle observation (odd population/sample size) or the average
of the middle two observations (even population/sample size) are
identified. We will denote $M$ as the population median and $m$ as the sample median.

**Example2.6: NHL BMI's and Rock and Roll Marathon Speeds**

Using the **length**, **sum**, **mean** and **median** functions in R, we obtain $N$, $\sum_{i=1}^Ny_i$, $\mu$, and the medians
for NHL BMI's in Table \@ref(tab:nhlbmi2) and marathon speeds by gender for the
Rock and Roll marathon in Table \@ref(tab:rrmarathon4). 

```{r nhlbmi2, echo=FALSE, tidy=FALSE}

nhl <- read.csv("nhl_ht_wt_1415.csv")
#head(nhl,2)

bmi.nhl <- 703*nhl$Weight/(nhl$Height^2)

### Use built-in length, sum,  mean, and median functions
bmi.nhl1.out <- cbind(length(bmi.nhl), sum(bmi.nhl),mean(bmi.nhl),
                    median(bmi.nhl))
colnames(bmi.nhl1.out) <- c("$N$", "$\\sum_iy_i$", "$\\mu$", "$M$")
rownames(bmi.nhl1.out) <- c("NHL BMI")
# round(bmi.nhl1.out, 3)

knitr::kable(bmi.nhl1.out, escape=FALSE, digits=3,
             caption='Population size, sum, mean, and median for NHL BMI',
             booktabs=TRUE)
```

```{r rrmarathon4, echo=FALSE, tidy=FALSE}

rr.mar <- read.csv("rocknroll_marathon_mf2015a.csv")
head(rr.mar,2)

mean.F <- mean(rr.mar$mph[rr.mar$Gender=="F"])
median.F <- median(rr.mar$mph[rr.mar$Gender=="F"])
sum.F <- sum(rr.mar$mph[rr.mar$Gender=="F"])
n.F <- length(rr.mar$mph[rr.mar$Gender=="F"])
mean.M <- mean(rr.mar$mph[rr.mar$Gender=="M"])
median.M <- median(rr.mar$mph[rr.mar$Gender=="M"])
sum.M <- sum(rr.mar$mph[rr.mar$Gender=="M"])
n.M <- length(rr.mar$mph[rr.mar$Gender=="M"])

rr.mar.mean.out.F <- cbind(n.F, sum.F, mean.F, median.F)
rr.mar.mean.out.M <- cbind(n.M, sum.M, mean.M, median.M)
rr.mar.mean.out <- rbind(rr.mar.mean.out.F, rr.mar.mean.out.M)
colnames(rr.mar.mean.out) <- c("$N$", "$\\sum_iy_i$", "$\\mu$", "$M$")
rownames(rr.mar.mean.out) <- c("Females", "Males")
#round(rr.mar.mean.out, 3)

knitr::kable(rr.mar.mean.out, escape=FALSE, digits=3,
             caption='Population size, sum, mean, and median velocity (mph) Rock and Roll Marathon
             by gender',
             booktabs=TRUE)

#tapply(rr.mar$mph, rr.mar$Gender, length)
#tapply(rr.mar$mph, rr.mar$Gender, sum)
#tapply(rr.mar$mph, rr.mar$Gender, mean)
#tapply(rr.mar$mph, rr.mar$Gender, median)
```

Note that the mean (26.514) and median (26.542) of NHL BMI values are very
close, as is expected for an (approximately) symmetric distribution.

For the marathon speeds, we use the obtain the means and medians
directly by gender.

The marathon velocity distributions are skewed-right, with a few very
fast runners in each gender. This causes the means (F=5.84, M=6.37) to
be larger than the medians (F=5.71, M=6.28).

$$ \nabla $$

**Example 2.7: James Short's Measurements of the Sun's Parallax**

The parallax is defined as (Merriam-Webster Dictionary):

> "the apparent displacement or the difference in apparent direction of
> an object as seen from two different points not on a straight line
> with the object. especially : the angular difference in direction of a
> celestial body as measured from two points on the earth's orbit."

James Short reported $n=158$ measurements of the parallax of the sun in
seconds of a degree [@short_1763],
also reported in [@stigler_1977].
The summary calculations are given in
Table \@ref(tab:shortparallax1). The true value has since been determined to be 8.798. 

A histogram of
the data, the true value, and sample mean, as well as a box plot of the
measurements are given in Figure \@ref(fig:shortparallax1).

```{r shortparallax1, fig.cap='James Short Measurements of the Parallax of the Sun', out.width='80%', fig.asp=.75, fig.align='center', tidy=FALSE, echo=FALSE}

short1 <- read.csv("short_parallaxsun.csv")
head(short1, 2)

short1.out <- cbind(length(short1$prlxsun), sum(short1$prlxsun),
mean(short1$prlxsun), median(short1$prlxsun),
mean(short1$prlxsun,trim=0.05))
colnames(short1.out) <- c("$n$", "$\\sum_iy_i$", "$\\bar{y}$", "$m$",
                          "90\\% Trim Mean")
rownames(short1.out) <- c("Parallax Measurements")
# round(short1.out, 4)

knitr::kable(short1.out, escape=F, digits=4,
             caption='Sample size, sum, sample mean, sample median, and 90 percent trimmed mean of sun parallax measurements',
             booktabs=TRUE)

par(mfrow=c(1,2))
hist(short1$prlxsun, breaks=seq(5.5,11,0.25), ylim=c(0,60),
     xlab="Sun Parallax", col="yellow",
  main="Histogram of Parallax")
abline(v=mean(short1$prlxsun), lty=1, lwd=2, col="red")
abline(v=8.798, lty=1, lwd=2, col="blue")
legend("topleft", c("Short's Mean", "TrueValue"), lty=c(1,1), lwd=c(2,2),
       col=c("red", "blue"))

boxplot(short1$prlxsun, main="Box Plot of Parallax",
       ylab="Sun Parallax", col="pink")
par(mfrow=c(1,1))

```

$$ \nabla $$

**Outliers** are observations that lie "far" away from the others. These
may be data that have been entered erroneously or just individual cases
that are quite different from others. As stated above, means can be
affected by outliers, while medians generally are not. A measure of the
mean that is not affected by outliers is the **trimmed mean**. This is
the mean of observations in the "middle" of the measurements. For
instance, a 90% trimmed mean is the mean of the middle 90% of the
ordered measurements (removing the smallest 5% and largest 5%).

Note that the Short parallax data has some extreme outliers in the box
plot. The 90% trimmed mean is 8.594 (see above output, where we have
trimmed 5% in each tail) which is not far from the sample mean as the
data are still fairly symmetric despite the outliers.

### Measures of Variability

Along with the "location" of a set of measurements, researchers are also
interested in their variability (aka dispersion). The **range** is the
distance between the largest and smallest measurements (note that this
differs from the standard meaning which would just give the lowest and
highest values). The **interquartile range** (IQR) is the distance
between the 75th percentile (3/4 of measurements lie below it) and the
25th percentile (1/4 of the measurements lie below it). That is, the IQR
measures the range for the middle half of the ordered measurements.

Measures that are more widely used in making inferences are the
**variance** and its square root, the **standard deviation**. In terms
of measurements, the variance is approximately the average squared
distance of the individual measurements from the mean (for a population,
it is the average). The formulas for the population and sample variance
are given below. Note that unless stated otherwise specifically,
software packages are reporting the sample version.

$$ \mbox{Population Variance: } \sigma^2 = \frac{\sum_{i=1}^N\left(y_i-\mu\right)^2}{N} \qquad \qquad
\mbox{Sample Variance: } s^2  = \frac{\sum_{i=1}^n\left(y_i-\overline{y}\right)^2}{n-1} $$

The reason for dividing by $n-1$ in the sample variance is to make the
estimator an unbiased estimator for the population variance. That is,
when computed across all possible samples, the "average" of the sample
variance will be the population variance. The standard deviation is the
positive square root of the variance and is in the same units as the
measurements. The population standard deviation is denoted as $\sigma$,
the sample standard deviation is denoted as $s$. For many (but certainly
not all) distributions, approximately 2/3 of the measurements lie within
one standard deviation of the mean and approximately 19/20 lie within
two standard deviations of the mean. 

**Example 2.8: NHL BMI's and Rock and Roll Marathon Speeds**

We compute the minimum, maximum, range, lower and upper quartiles, and the interquartile range, for NHL BMI's in Table \@ref(tab:nhlbmi3).

The means, variances, and standard
deviations for the NHL BMI's and the Rock and Roll marathon speeds by
gender are given in Table \@ref(tab:nhlbmi3a), 
and Table \@ref(tab:rrmarathon5), respectively. 
Since we treat each of these as a population, we will make a
slight adjustment to R's "built-in" functions **var** and **sd**, which
compute the sample versions by default. Also included in the the
results are the proportions of measurements lying within 1 and 2 standard deviations of the mean.

```{r nhlbmi3, echo=FALSE, tidy=FALSE}

nhl <- read.csv("nhl_ht_wt_1415.csv")
#head(nhl, 2)

bmi.nhl <- 703*nhl$Weight/(nhl$Height^2)

## Obtain Summary Statistics 
bmi.max <- max(bmi.nhl)    # Highest BMI
bmi.min <- min(bmi.nhl)    # Lowest BMI
range <- bmi.max - bmi.min # Compute Range
bmi.75 <- quantile(bmi.nhl,.75)    # BMI 75%-ile
bmi.25 <- quantile(bmi.nhl,.25)    # BMI 25%-ile
IQR <-  IQR(bmi.nhl)    # Compute IQR
N <- length(bmi.nhl)     # Use "length" function to get N
mu <- mean(bmi.nhl)      # Use "mean" function to get mu
sum.dev2 <- sum((bmi.nhl - mu)^2) # Numerator of Variance
sigma2 <- sum.dev2/N     # Population Variance
s2 <- sum.dev2/(N-1)     # Sample Variance
sigma <- sqrt(sigma2)    # Population Standard Deviation
s <- sqrt(s2)            # Sample Standard Deviation
s2.a <- var(bmi.nhl)     # Sample Variance with "var" function
sigma2.a <- (N-1)*var(bmi.nhl)/N  # Pop variance with "var" function
s.a <- sd(bmi.nhl)           # Sample Std Dev with "sd" function
sigma.a <- sqrt((N-1)/N)*sd(bmi.nhl)  # Population Std Dev with "sd" function
## Proportion of Individual w/in 1 and 2 SDs of mean
mu_pm_1s <- sum(bmi.nhl >= mu-sigma & bmi.nhl <= mu+sigma) / N
mu_pm_2s <- sum(bmi.nhl >= mu-2*sigma & bmi.nhl <= mu+2*sigma) / N

## Generate Summary Table
bmi.var.out1 <- cbind(bmi.min,bmi.max,range,bmi.25,bmi.75,IQR)
colnames(bmi.var.out1) <- c("min","max","range","LQ","UQ","IQR")
rownames(bmi.var.out1) <- c("NHL BMI")
# round(bmi.var.out1, 3)

knitr::kable(bmi.var.out1, digits=3,
      caption='Minimum, Maximum, Range, Lower and Upper Quartiles for NHL BMI',
      booktabs=TRUE)
```

```{r nhlbmi3a, echo=FALSE, tidy=FALSE}
bmi.var.out2 <- cbind(mu,sum.dev2,sigma2,sigma,mu_pm_1s,mu_pm_2s)
colnames(bmi.var.out2) <- c("$\\mu$","$\\sum(y-\\mu)^2$","$\\sigma^2$","$\\sigma$",
                          "$P(\\mu\\pm\\sigma)$","$P(\\mu\\pm2\\sigma)$")
rownames(bmi.var.out2) <- c("NHL BMI")
# round(bmi.var.out2, 3)

knitr::kable(bmi.var.out2, digits=3, escape=FALSE,
      caption='Mean, Sum of Squares, Variance, SD, Proportions within 1 and 2 SDs of mean for
      NHL BMI',
      booktabs=TRUE)
```

For the marathon speeds, we will simply use the **var** and **sd**
functions in R, applied separately to Females and Males. As both
population sizes exceed 1000, the adjustment for population variances
and standard deviations would be very small.

```{r rrmarathon5, echo=FALSE, tidy=FALSE}

rr.mar <- read.csv("rocknroll_marathon_mf2015a.csv")
head(rr.mar,2)

## Obtain mean and standard deviation by gender
#tapply(rr.mar$mph,rr.mar$Gender,mean)
#tapply(rr.mar$mph,rr.mar$Gender,median)

#tapply(rr.mar$mph,rr.mar$Gender,var)
#tapply(rr.mar$mph,rr.mar$Gender,sd)

## Get summary statistics and table

N.M <- length(rr.mar$mph[rr.mar$Gender=="M"])
N.F <- length(rr.mar$mph[rr.mar$Gender=="F"])
mu.M <- mean(rr.mar$mph[rr.mar$Gender=="M"])
mu.F <- mean(rr.mar$mph[rr.mar$Gender=="F"])
sigma2.M <- var(rr.mar$mph[rr.mar$Gender=="M"])
sigma2.F <- var(rr.mar$mph[rr.mar$Gender=="F"])
sigma.M <- sqrt(sigma2.M)
sigma.F <- sqrt(sigma2.F)

mu_pm_1s.M <- sum(rr.mar$mph[rr.mar$Gender=="M"] >= mu.M-1*sigma.M & 
                  rr.mar$mph[rr.mar$Gender=="M"] <= mu.M+1*sigma.M) / N.M
mu_pm_2s.M <- sum(rr.mar$mph[rr.mar$Gender=="M"] >= mu.M-2*sigma.M & 
                  rr.mar$mph[rr.mar$Gender=="M"] <= mu.M+2*sigma.M) / N.M
mu_pm_1s.F <- sum(rr.mar$mph[rr.mar$Gender=="F"] >= mu.F-1*sigma.F & 
                  rr.mar$mph[rr.mar$Gender=="F"] <= mu.F+1*sigma.F) / N.F
mu_pm_2s.F <- sum(rr.mar$mph[rr.mar$Gender=="F"] >= mu.F-2*sigma.F & 
                  rr.mar$mph[rr.mar$Gender=="F"] <= mu.F+2*sigma.F) / N.F

rr.var.out.M <- cbind(mu.M,sigma2.M,sigma.M,mu_pm_1s.M,mu_pm_2s.M)
rr.var.out.F <- cbind(mu.F,sigma2.F,sigma.F,mu_pm_1s.F,mu_pm_2s.F)
rr.var.out <- rbind(rr.var.out.F,rr.var.out.M)
colnames(rr.var.out) <- c("$\\mu$","$\\sigma^2$","$\\sigma$",
                          "$P(\\mu \\pm \\sigma)$","$P(\\mu \\pm  2\\sigma)$")
rownames(rr.var.out) <- c("Females","Males")
#round(rr.var.out,3)

knitr::kable(rr.var.out, escape=F, digits=3,
             caption='Means, Variances, SDs, Proportions within 1 and 2 SD of mean for Rock and Roll marathon by gender',
             booktabs=TRUE)
```

Male speeds tend to be higher and more variable than Female speeds. All
three distributions have approximately 2/3 of individuals lying with one
standard deviation of the mean, and approximately 95% lying within two
standard deviations from the mean.

$$ \nabla $$

**Example 2.9: James Short's Measurements of the Sun's Parallax**

We compute the range, interquartile range, variance, and standard
deviations for the sample of $n=158$ sun parallax measurements in 
Table \@ref(tab:shortparallax2).

```{r shortparallax2, echo=FALSE, tidy=FALSE}

short1 <- read.csv("short_parallaxsun.csv")
head(short1, 2)

short2.out <- cbind(
length(short1$prlxsun),
mean(short1$prlxsun),
max(short1$prlxsun) - min(short1$prlxsun),
IQR(short1$prlxsun),
var <- var(short1$prlxsun),
sd <- sd(short1$prlxsun),
mean(short1$prlxsun >= mean(short1$prlxsun)-sd & short1$prlxsun <= mean(short1$prlxsun)+sd),
mean(short1$prlxsun >= mean(short1$prlxsun)-2*sd & short1$prlxsun <= mean(short1$prlxsun)+2*sd))

colnames(short2.out) <- c("$n$","$\\bar{y}$", "range","IQR", "$s^2$","$s$",
                          "$P(\\bar{y} \\pm s)$","$P(\\bar{y} \\pm 2s)$")
rownames(short2.out) <- c("Parallax")
# round(short2.out,3)

knitr::kable(short2.out, escape=F, digits=3,
             caption='Sample size, mean, range, IQR, variance, SD, and proportions within 1 and 2 standard deviations of the mean, sun parallax data',
             booktabs=TRUE)

```

The full set of measurements lie within a range of 5.04 seconds of a
degree, while the middle 50% lie within a range of 0.445. The variance
is 0.455 and the standard deviation (a typical distance from an
observation to the mean) is 0.674. Further, approximately 77.8% of
measurements lie within one standard deviation and 93.7% lie within two
standard deviations of the mean.

$$ \nabla $$

## Describing More than One Variable

So far, we have looked at cases one variable at a time, although the
marathon speed data set has two variables: speed and gender. Now we
consider describing relationships when two variables are observed on
each sampling/experimental unit. These can be extended to more than two
variables, but can be harder to visualize. We consider graphical
techniques as well as numerical measures. Keep in mind that variable
types (nominal, ordinal, and numeric) will dictate which method(s) is
(are) appropriate.

When both variables are categorical (nominal or ordinal), two methods of
plotting them are **stacked bar graphs** and **cluster bar graphs**. For
the stacked bar graph, one variable is on the horizontal axis (one slot
for each level) and the other variable is displayed within the bars with
subcategories for each of its levels. In a cluster (grouped) bar graph,
one variable forms "major groupings," while the second variable is
plotted "side-by-side" within the groupings. Both methods are based on
results of a **contingency table** also known as a **crosstabulation**.
These are tables where rows are the levels of one categorical variable,
columns are levels of another variable, and numbers within the table are
counts of the number of units falling in that cell (combination of
variable levels). Often these are converted into proportions either
overall (cell probabilities sum to 1), or within rows or columns
marginally.

**Example 2.10: Thumb Styles of Blues Guitarists by Region and Period**

A study reported hand and thumb styles of Blues guitarists as well as
the region they were from and when they were born [@cohen_1996]. The regions are 1=East, 2=Delta, and 3=Texas. The thumb
styles are 1=Alternating, 2=Utility, and 3=Dead. The birth period was
labeled post1906 with 0=Born before 1906, 1=Born after 1906. First, the
association between region (row) and thumb style (column) is considered,
then birth period is added. The crosstabulation is given in Table \@ref(tab:bluesguitar1a).
The marginal counts by region and thumb style are given in Table \@ref(tab:bluesguitar1b)
and Table \@ref(tab:bluesguitar1c), respectively. The joint proportions are
given in Table \@ref(tab:bluesguitar1d). The row proportions of thumb style within region
are given in Table \@ref(tab:bluesguitar1e). The column proportions of region within
thumb style are given in Table \@ref(tab:bluesguitar1f).
Figure \@ref(fig:bluesguitar2) gives the Stacked and Cluster Bar Graphs.
 
```{r bluesguitar1a, echo=FALSE, tidy=FALSE}
blues <- read.csv("blues_hand.csv")
blues$region <- factor(blues$region, levels=1:3, labels=c("East", "Delta", "Texas"))
blues$thumbSty <- factor(blues$thumbSty, levels=1:3, labels=c("Alternating", "Utility", "Dead"))
reg_ts <- table(blues$region, blues$thumbSty)
rownames(reg_ts) <- c("East", "Delta", "Texas")
colnames(reg_ts) <- c("Alternating", "Utility", "Dead")

knitr::kable(reg_ts,
             caption='Crosstabulation of blues guitarists by region and thumb style',
             booktabs=TRUE)
```
```{r bluesguitar1b, echo=FALSE, tidy=FALSE}
# Obtain Row (1) and Column (2) Marginal Totals

knitr::kable(margin.table(reg_ts,1), caption='Marginal counts of blues guitarists by region',
  booktabs=TRUE)
```

```{r bluesguitar1c, echo=FALSE, tidy=FALSE}

knitr::kable(margin.table(reg_ts,2), caption='Marginal counts of blues guitarists by thumb style',
  booktabs=TRUE)
```
```{r bluesguitar1d, echo=FALSE, tidy=FALSE}
# Obtain Proportions across all Cells

knitr::kable(reg_ts/sum(reg_ts), digits=4, caption='Joint proportions of blues guitarists by region and thumb style',
  booktabs=TRUE)
```
```{r bluesguitar1e, echo=FALSE, tidy=FALSE}
# Obtain Row Proportions (Thumb Style w/in Region)

knitr::kable(prop.table(reg_ts,1), digits=4,
  caption='Proportions by thumb style within region',
   booktabs=TRUE)
```

```{r bluesguitar1f, echo=FALSE, tidy=FALSE}
# Obtain Column Proportions (Region w/in Thumb Style)

knitr::kable(prop.table(reg_ts,2), digits=4,
  caption='Proportions by region within thumb style',
  booktabs=TRUE)
```

```{r bluesguitar2, fig.cap='Stacked and Cluster Bar Charts for Blues Guitarists', out.width='80%', fig.asp=.75, fig.align='center', tidy=FALSE, echo=FALSE}
## Obtain Cluster (Grouped) and Stacked Bar Plots
## t(prop.table(reg_ts,1)) takes transpose so that group var is Region

## Figure
par(mfrow=c(1,2))   # 1 row and 2 columns of plots
barplot(t(prop.table(reg_ts,1)),beside=T,
    legend=colnames(reg_ts),ylim=c(0,1), col=c("coral", "orange", "yellow"),
    main="Grouped Bar Plot - Blues Guitarists")

barplot(t(prop.table(reg_ts,1)),beside=F,
   legend=colnames(reg_ts), col=c("coral", "orange", "yellow"),
   main="Stacked Bar Plot - Blues Guitarists", ylim=c(0,1.40))

par(mfrow=c(1,1))
```

$$ \nabla $$
**Example 2.11: IMDB User Reviews for 14 Disaster Films**

The Internet Movie Database has user reviews for a huge number of films. In this example, we
consider the proportion of reviews (from 1-10) for 14 disaster films (Independence Day (I), The Core, Armageddon, Deep Impact, Geostorm, San Andreas, Independence Day (II), 2012, Day After Tomorrow, War of the Worlds, Poseidon, Perfect Storm, Twister (I), and Titanic). A table of the proportions by rating score for the 14 films is given in Table \@ref(tab:disastermovie1). A grouped bar chart is
given in Figure \@ref(fig:disastermovie1) and a stacked bar chart is given in Figure \@ref(fig:disastermovie2).

```{r disastermovie1, fig.cap='Stacked Bar Chart for user reviews (1-10) of 14 Disaster Movies on IMDB', out.width='80%', fig.asp=.75, fig.align='center', echo=FALSE, tidy=FALSE}
movie <- read.csv("movie_disaster.csv")
head(movie,2)
movie1 <- movie[,-1]

movie.table <- matrix(movie1$numRate,ncol=10)
colnames(movie.table) <- unique(movie1$Rating)
rownames(movie.table) <- unique(movie1$Movie1)
rate.movie.prop <- prop.table(movie.table,1)

knitr::kable(rate.movie.prop, digits=3,
             caption='IMDB user ratings for 14 disaster films',
             booktabs=TRUE)

color3 <- brewer.pal(ncol(movie.table), "Set3")

barplot(t(prop.table(movie.table,1)),beside=T, cex.names=0.5,
    ylim=c(0,0.4), col=color3,
    main="Grouped Bar Plot - Disaster Movies")
```

```{r disastermovie2, fig.cap='Cluster Bar Chart for user reviews (1-10) of 14 Disaster Movies on IMDB', out.width='80%', fig.asp=.75, fig.align='center', echo=FALSE}
barplot(t(prop.table(movie.table,1)),beside=F, cex.names=0.5,
   col=color3,
   main="Stacked Bar Plot - Disaster Movies", ylim=c(0,1.0))
```

$$\nabla $$

When the independent variable is categorical (nominal or ordinal) and
the response (dependent variable) is numeric, we can construct
side-by-side histograms and density plots, or box plots (see Figure
\@ref(fig:michelson2) for side-by-side box plots). Histograms and
densities can also be placed into single plots with different colors or
patterns as in Figure \@ref(fig:rrmarathon6).

```{r rrmarathon6, fig.cap='Density Plots of Velocities at Rock and Roll Marathon 2015 (mph)', out.width='80%', fig.asp=.75, fig.align='center', echo=FALSE}

rr.mar <- read.csv("rocknroll_marathon_mf2015a.csv")
head(rr.mar,2)

p <- ggplot(rr.mar, aes(x=mph)) +   
  geom_density(aes(group=Gender, color=Gender, fill=Gender), adjust=4, alpha=1/2) +
    scale_color_discrete() + scale_fill_discrete() + scale_fill_manual(values=c("green", "purple")) +
  ggtitle("MPH Densities by Gender")

p

```

When two variables (labeled $x$ and $y$) are both numeric, one numeric
descriptive measure that is widely reported is the **correlation**
between the two variables. Technically, this is called the Pearson
product moment coefficient of correlation. This measure is only for the
**linear**, or "straight line" relation between the two variables.
Unlike in Regression (described later), the variables are not
necessarily (but can be) identified as an independent and or dependent
variable. The formula for this measure (population and sample) are given
below.

$$ \mbox{Population Correlation: } \rho = \frac{
\frac{1}{N}\sum_{i=1}^N\left(x_i-\mu_x\right)\left(y_i-\mu_y\right)}{\sigma_x\sigma_y}=
\frac{\sum_{i=1}^N\left(x_i-\mu_x\right)\left(y_i-\mu_y\right)}{\sqrt{\sum_{i=1}^N\left(x_i-\mu_x\right)^2\sum_{i=1}^N\left(y_i-\mu_y\right)^2}} $$

$$ \mbox{Sample Correlation: } r =  \frac{\frac{1}{n-1}\sum_{i=1}^n\left(x_i-\overline{x}\right)\left(y_i-\overline{y}\right)}{s_xs_y}=
\frac{\sum_{i=1}^n\left(x_i-\overline{x}\right)\left(y_i-\overline{y}\right)}{\sqrt{\sum_{i=1}^n\left(x_i-\overline{x}\right)^2\sum_{i=1}^n\left(y_i-\overline{y}\right)^2}} $$

A **scatterplot** is a plot where each case's $x$ and $y$ pairs are
plotted in two dimensions. When one variable is the dependent variable,
it is labeled $y$, and plotted on the vertical axis and the independent
variable is labeled $x$, plotted on the horizontal axis. We are
interested in any pattern (linear or possibly nonlinear, or none at all)
between the variables. The formulas for the (ordinary) least squares
regression line relating $y$ to $x$ are given below.

$$ \hat{y} = \hat{\beta}_0 + \hat{\beta}_1 x \qquad \qquad \hat{\beta}_1 = 
\frac{\sum_{i=1}^n\left(x_i-\overline{x}\right)\left(y_i-\overline{y}\right)}{\sum_{i=1}^n\left(x_i-\overline{x}\right)^2}
 \qquad \qquad \hat{\beta}_0 = \overline{y} - \hat{\beta}_1 \overline{x} $$

$$SSE=\sum_{i=1}^n\left(y_i-\hat{y}_i\right)^2
= \sum_{i=1}^n\left(y_i-\left(\hat{\beta}_0 + \hat{\beta}_1 x_i\right)\right)^2 $$

**Example 2.12: Relation Between Temperature and Water Evaporation**

An experiment was conducted that observed the temperature $x$
(fahrenheit) and water evaporation $y$ (grains of water) with
measurements taken at 8:00AM daily from 11/10/1692-11/09/1693 [@halley_1694].

The plot of the data and the linear regression equation is given in
Figure \@ref(fig:halley1). The correlation and regression equation were
obtained using the **cor** and **lm** functions.

```{r halley1, fig.cap='Temperature (X) and Water Evaporation (Y) - Edmund Halley Observations 1692/1693', out.width='80%', fig.asp=.75, fig.align='center'}

evap <- read.table("evap.dat",
        header=F, 
        col.names=c("day", "dayEvap", "dayTemp", "dayPres", "mdate"))
head(evap,2)

## Compute correlation coefficient (Pearson) between evaporation and temp
cor(evap$dayEvap, evap$dayTemp)

## Fit simple linear regression model
mod1 <- lm(dayEvap ~ dayTemp, data=evap)
summary(mod1)
anova(mod1)

plot(evap$dayEvap ~ evap$dayTemp, pch=16, col="red", cex=0.7,
    main="Daily Water Evaporation (y) and Temperature (x)",
    ylab="Water Evaporation", xlab="Temperature")
abline(mod1, col="blue", lwd=2)
```

The sample correlation is $r =0.7691$ and the fitted linear regression
equation is $\hat{y}=23.9456+0.6288x$.

$$ \nabla  $$

**Example 2.13: Heights of Adult Children and Their Parents**

Francis Galton measured many aspects of humans, plants, and animals
during the late 1800s, some of which were presented in table form in his
book *Natural Inheritance*. One analysis that had been published
previously [@galton_1886]
introduced the notion of
linear regression. Galton reported the heights of adult children and
their
"mid-parents" which was the average height of the parents. Galton multiplied female heights for the adult children and the mothers by 1.08 to make the female and male heights"comparable."
The individual data were obtained from Galton's notebooks and are
available due to Professor James A. Hanley [@hanley_2004].

Histograms of the male and (unscaled) female heights is given in Figure
\@ref(fig:galton1). The histograms are approximately mound-shaped within
gender. The plot of adult child height versus mid-parent height (with
female heights scaled by 1.08) is given in Figure \@ref(fig:galton2).
The plot contains three lines, which are described below.

```{r galton1, fig.cap='Histograms of Adult Male and Female Children Heights - Frances Galton Measurements', out.width='80%', fig.asp=.75, fig.align='center', echo=FALSE}

galton <- read.csv("Galton_Data.csv")
head(galton,2)

mod1 <- lm(AdltChld ~ MidParent, data=galton)
summary(mod1)

par(mfrow=c(2,1))
hist(galton$Height[galton$Gender == "M"], breaks=seq(56,80,1.0), freq=F, col="purple", 
   main="Male Adult Child Heights", xlab="Height (inches)")
lines(density(galton$Height[galton$Gender == "M"]))

hist(galton$Height[galton$Gender == "F"], breaks=seq(56,80,1.0), freq=F, col="green",
   main="Female Adult Child Heights", xlab="Height (inches)")
lines(density(galton$Height[galton$Gender == "F"]))
par(mfrow=c(1,1))

```

```{r galton2, fig.cap='Adult Chidren Scaled Heights (Y) versus Midparents Height (X) and Regression Line', out.width='80%', fig.asp=.75, fig.align='center', echo=FALSE}

par(mfrow=c(1,1))
plot(jitter(galton$AdltChld) ~ jitter(galton$MidParent), pch=16, cex=0.6,
    xlab="Mid Parent Height", ylab="Adult Child Height",
    main="Height of Adult Child vs Mid Parent Height")
abline(lm(galton$AdltChld ~ galton$MidParent), lwd=2.5, col="red")
abline(a=0,b=1, col="darkblue", lwd=2.5, lty=1)
abline(h=mean(galton$MidParent), col="darkgreen", lwd=2.5, lty=1)
legend("topleft", c("Y-hat = 18.77+0.73X", "Y-hat=X", "Y-hat=69.22"),
      col=c("red", "darkblue", "darkgreen"), 
      lwd=c(2.5,2.5,2.5), lty=1)

```

-   Steepest: Line of equality $\hat{y}=x$, which represents the case with the
    average adult child height equaling the mid-parent height.
-   Flat: Constant line $\hat{y}=\overline{y}$, which represents the case with
    the average adult child height equaling the average mid-parent
    height (no association between adult child and mid-parent height).
-   Middle: Least squares regression line $\hat{y} = 18.77 + 0.73x$

The fact the least squares line falls between the two reference lines
showed that adult children of tall parents tended to be tall, but not as
tall on average as their parents. Similarly, adult children of short
parents tended to be short, but not as short on average as their
parents. Galton referred to this phenomenon as "regression to
mediocrity." Today it is more widely referred to as "regression to the
mean."

$$ \nabla  $$

We often are interested in relationships among more than two numeric
variables. Scatterplot and correlation matrices can be constructed to
demonstrate the bivariate association of all pairs of variables.


**Example 2.14: Compressive Strength and Microfabric Properties of Amphibolites**

A study reported the relationship between Uniaxial Compression Strentgh (UCS) and 8 predictor variables including: percent hornblende (hb), grain size (gs), and grain area (ga) [@ali_2014]. 
A simple scatterplot matrix of plots of all pairs of these
four variables is given in Figure \@ref(fig:rockstrength). The
correlation matrix is given along with R code below. Note that this can
be extended to all pairs of variables, the plot just gets very difficult
to focus on particular pairs of variables.

```{r rockstrength, fig.cap='Scatterplot Matrix for all pairs among 4 rock strength variables', out.width='80%', fig.asp=.75, fig.align='center'}
rs1 <- read.csv("rockstrength.csv")
head(rs1,2)

## Scatterplot matrix of UCS, hb, gs, ga (columns 2,6,7,8 of rs1)
plot(rs1[,c(2,6,7,8)])

## Obtain correlation matrix of UCS, hb, gs, ga  (columns 2,6,7,8 of rs1)
cor(rs1[,c(2,6,7,8)])

```

$$ \nabla $$

When data are highly skewed, individual cases have the ability to have a
large impact on the correlation coefficient. An alternative measure that
is widely used is the Spearman Rank Correlation Coefficient (aka
Spearman's rho). This coefficient is computed by ranking the $x$ and $y$
values from 1 (smallest) to $n$ or $N$ (largest), and applying the
formula for Pearson's coefficient to the ranks. This way, extreme $x$ or
$y$ values do not have as large of an impact on the coefficient. Also,
in many situations, the natural measurements are the rankings or
ordering themselves.

**Example 2.15: NASCAR Start and Finish Positions 1975-2003**

A study of NASCAR races for the years 1975-2003, considered the
correlation between starting and finishing positions among drivers for
the 898 races during those seasons 
[@winner_2006]. As
the data were orderings, it was natural to compute the correlation using
Spearman's rank correlation. The summary of the correlations is given
below, and a density plot and histogram are given in Figure
\@ref(fig:nasrace1).

```{r nasrace1, fig.cap='NASCAR Races 1975-2003 - Spearman rank correlation coefficient for start/finish positions', out.width='80%', fig.asp=.75, fig.align='center', echo=FALSE}
nasRace <- read.fwf("nascarr.dat",
    widths=c(3,6,4,4,9,7,9,9,7,5,3,4,4,9,7,8,5,38),
    col.names=c("seriesRace","year","yearRace","numCar","payout","cpiU",
       "spearman","kendall","trkLength","lapsComp","roadRace",
       "cautionFlag","leadChange","winTime","trkLat","trkLong",
       "trkCode","trkName"))

#length(nasRace$spearman)
#summary(nasRace$spearman)

hist(nasRace$spearman,breaks=seq(-.5,1,.02),prob=T, col="coral",
   xlab=expression(paste("Spearman's ",rho)), 
   main=expression(paste("NASCAR Start/Finish - Spearman's ", rho)))
lines(density(nasRace$spearman))

```

$$ \nabla $$

Many series (particularly when measured over time) display 
**spurious correlations**, 
particularly when both variables tend to increase or
decrease together with no **causal** reason that the two (or more)
variables move in tandem. For instance, the correlation between annual
U.S. internet users (per 100 people) and electrical power consumption
(kWh per capita) for the years 1994-2010 is .7821 (data source: The
World Bank). Presumably increasing internet usage isn't leading to large
increases in electrical consumption, or vice versa.
