---
output:
  pdf_document: default
  html_document: default
---
```{r}
library(tidyverse)
library(kableExtra)
```


# Experimental Design and the Analysis of Variance

Chapter 6 covered methods to make comparisons
between the means of a numeric response variable for two treatments or groups.
The cases were considered where the experiment was conducted as an independent samples (aka parallel groups, between subjects)
design, as well as a paired (aka crossover, within subjects) design. 

This chapter will introduce methods that can be used to compare
more than two groups (that is, when the explanatory variable has more
than two levels). In this chapter, we will refer to explanatory variable
as a **factor**, and their levels as **treatments**. The following situations will be covered.


- 1-Treatment Factor, Independent Samples Designs (Completely Randomized Design)
- 1-Treatment Factor, Paired Designs (Randomized Block Design)
- 2-Treatment Factors, Independent Samples Designs (Factorial Design)


In all situations, there will be a numeric response variable, and at 
least one categorical (or possibly numeric, with several levels)
independent variable. The goal will always be to 
compare mean (or median) responses among several populations.
When all factor levels for a factor are included in the experiment,
the factor is said to be **fixed**. When a sample of a larger population of
factor levels are included, the factor is said to be **random**. Only fixed effects designs are considered here.

## Completely Randomized Design (CRD) For Independent Samples
 
In the Completely Randomized Design, there is one factor that is
controlled. This factor has $k$ levels
(which are often treatment groups), and $n_i$ units are measured for
the $i^{th}$ level of the factor. 
Observed responses are defined as $y_{ij}$, representing the
measurement on the $j^{th}$ experimental unit (subject), receiving the $i^{th}$
treatment. We will write this in model form based on random responses as follows where the factor is **fixed**
(all levels of interest are included in the experiment).

$$ Y_{ij} = \mu + \alpha_i + \epsilon_{ij} = \mu_i + \epsilon_{ij}   \quad i=1,\ldots ,k; \quad j=1,\ldots ,n_i $$

Here, $\mu$ is the overall mean measurement across all treatments,
$\alpha_i$ is the effect of the $i^{th}$ treatment ($\mu_i=
\mu + \alpha_i$), and $\epsilon_{ij}$ is a
random error component that has
mean 0 and variance $\sigma^2$. This $\epsilon_{ij}$ allows for 
the fact that there will be variation among the measurements of
different subjects (units) receiving the same treatment. A common parameterization that has nice properties is to 
assume $\sum\alpha_i = 0$.

Of interest to the experimenter is whether or not there is a
**treatment effect**, that is do any of the levels of the treatment provide
higher (lower) mean response than other levels. This can be hypothesized
symbolically as follows.

$$H_0: \alpha_1 = \alpha_2 = \cdots = \alpha_k = 0 \mbox{ versus } H_A: \mbox{ Not all }
\alpha_i = 0  $$

The null hypothesis states that all $k$ means are equal, while the alternative states that they are not all equal.
Note that if $\alpha_1 = \alpha_2 = \cdots = \alpha_k = 0$
then $\mu_1=\cdots=\mu_k$.

As with the case where there are 
two treatments to compare, tests based on the assumption that
the $k$ populations are normal (mound-shaped) will be used, either assuming equal or unequal variances. Also, alternative tests (based
on ranks) that do not assume that the $k$ populations are
normal can be used to compare population medians. These are the **Kruskal-Wallis Test** for the CRD and **Friedman's Test** for the RBD. 
These tests will not be covered in these notes. 


### Analysis of Variance and $F$-Test for Normal Data

When the underlying populations of measurements that are to be compared
are approximately normal, with equal variances, the $F$-test is appropriate. To conduct this test,
partition the total variation in the sample data to variation **within**
and **among** treatments. This partitioning is referred to as the
**Analysis of Variance** and is an important tool in many statistical
procedures. First, we define the following items, based on random outcomes $Y_{ij}$ where $i$ indexes treatment and $j$ represents
the replicate number, with $n_i$ observations for treatment $i$ and $N=n_1+\cdots + n_k$. These items include various means, sums of squares (SS),
degrees of freedom (df), and mean squares (MS).

$$ Y_{ij} \sim N\left(\mu_i, \sigma\right)  $$

$$ \overline{Y}_{i.}=\frac{\sum_{j=1}^{n_i}Y_{ij}}{n_i} \qquad
\overline{Y}_{i.} \sim N\left(\mu_i, \frac{\sigma}{\sqrt{n_i}}\right)  $$

$$ \overline{Y}_{..}=\frac{\sum_{i=1}^k\sum_{j=1}^{n_i}Y_{ij}}{N} = \frac{\sum_{i=1}^k n_i \overline{Y}_{i.}}{N}  $$

$$ \mbox{Total (Corrected) Sum of Squares: } TSS=\sum_{i=1}^k\sum_{j=1}^{n_i}\left(Y_{ij}-\overline{Y}_{..}\right)^2 
 \qquad df_{\mbox{Tot}}=N-1 $$

$$ \mbox{Between Treatment Sum of Squares: } SST=\sum_{i=1}^k\sum_{j=1}^{n_i}\left(\overline{Y}_{i.}-\overline{Y}_{..}\right)^2
=\sum_{i=1}^k n_i\left(\overline{Y}_{i.}-\overline{Y}_{..}\right)^2 $$
$$df_{T}=k-1 \qquad \qquad MST=\frac{SST}{k-1} $$ 

$$ \mbox{Within Treatment (Error) Sum of Squares: } SSE=\sum_{i=1}^k\sum_{j=1}^{n_i}\left(Y_{ij}-\overline{Y}_{i .}\right)^2 
=\sum_{i=1}^k \left(n_i-1\right)S_i^2 $$
$$ df_{E}=N-k \qquad \qquad MSE=\frac{SSE}{N-k}$$

Under the null hypothesis of no treatment effects $(\mu_1=\cdots = \mu_k=\mu)$, or equivalently $(\alpha_1=\cdots = \alpha_k=0)$
the following results are obtained, where $MST$ and $MSE$ are mean squares for treatments and error, respectively.

$$ E\left\{MST\right\}=E\left\{\frac{SST}{k-1}\right\}=\sigma^2  \qquad
 E\left\{MSE\right\}=E\left\{\frac{SSE}{N-k}\right\}=\sigma^2  $$

Under the null hypothesis of no treatment effects, $E\left\{MST\right\}=E\left\{MSE\right\}=\sigma^2$ and the ratio $MST/MSE$
follows the $F$-distribution with $k-1$ numerator and $N-k$ denominator degrees of freedom. When the null is not true and
not all $\alpha_i=0$, then the ratio follows the non-central $F$-distribution with parameter $\lambda$ given below.

$$ \frac{MST}{MSE} \sim F_{\nu_1,\nu_2,\lambda} \qquad \lambda=
 \frac{\sum_{i=1}^k n_i \alpha_i^2}{\sigma^2} \qquad \nu_1=k-1 \qquad \nu_2=N-k $$

Once samples have been obtained and the $y_{ij}$ are observed, the $F$-test is conducted as follows.

$$\overline{y}_{i.} = \frac{\sum_{j=1}^{n_i}y_{ij}}{n_i} \qquad
s_i = \sqrt{\frac{\sum_{j=1}^{n_i} (y_{ij}-\overline{y}_{i.})^2}{n_i-1}} \qquad
N =n_1 + \cdots + n_k \qquad
\overline{y}_{..} = \frac{\sum_{i=1}^k \sum_{j=1}^{n_i}y_{ij}}{N} $$
$$TSS=\sum_{i=1}^k \sum_{j=1}^{n_i} \left(y_{ij} - \overline{y}_{..} \right)^2 \qquad
SST =\sum_{i=1}^k n_i\left(\overline{y}_{i.} - \overline{y}_{..} \right)^2 \qquad 
SSE = \sum_{i=1}^k (n_i -1) s_i^2    $$


Here, $\overline{y}_{i.}$ and $s_i$ are the mean and standard deviation of measurements
in the $i^{th}$ treatment group, and $\overline{y}_{..}$ and $N$ are the overall mean and
total number of all measurements. $TSS$ is the total variability
in the data (ignoring treatments), $SST$ measures the variability in 
the sample means among the treatments (weighted by the sample sizes), and $SSE$ measures the variability
within the treatments.

Note that the goal is to determine whether or not the population means
differ. If they do, we would expect $SST$ to be large, since that sum of
squares is measuring differences in the sample means. A test for treatment effects
is conducted after constructing an Analysis of
Variance table, as shown in Table \@ref(tab:crdaov1). In that table, there are
**sums of squares** for treatments ($SST$), for error ($SSE$), and total
($TSS$). Also, there are **degrees of freedom**, which represent
the number of "independent" terms in the sum of squares. Then,
the **mean squares**, are sums of squares divided by their
degrees of freedom. Finally, the $F$--statistic is computed as $F=MST/MSE$.
This will serve as the test statistic. Note that $MSE$ is an extension of
the pooled variance computed in Chapter 6 for two groups, and often
it is written as $MSE=s^2$.


```{r crdaov1, echo=FALSE, tidy=FALSE}
aovtable.out <- rbind(
cbind("Treatments", "k-1", "SST", "MST=SST/(k-1)", "F=MST/MSE"),
cbind("Error", "N-k", "SSE", "MSE=SSE/(N-k)", "   "),
cbind("Total", "N-1", "TSS", "   ","   ")
)
colnames(aovtable.out) <- c("Source", "Degrees of Freedom", "Sum of Squares", "Mean Square", "$F_{obs}$")

knitr::kable(aovtable.out, escape=FALSE, caption='The Analysis of Variance Table for the Completely Randomized
(Parallel Groups) Design',
   booktabs=TRUE)

```


The formal method of testing this hypothesis is as follows.


- $H_0: \alpha _1 = \cdots = \alpha_k = 0  \quad (\mu_1 = \cdots = \mu_k)$
 (No treatment effect)
- $H_A: \mbox{ Not all } \alpha_i$ are 0 (Treatment effects exist)
- T.S. $F_{obs} = \frac{MST}{MSE}$
- R.R.: $F_{obs} \geq F_{\alpha, k-1,N-k}$ 
- p-value: $P=P(F_{k-1, N-k} \geq F_{obs})$
 



**Example 7.1: Body Mass Indices of NHL, NBA, and EPL, Players**

Consider an extension of the Body Mass Index analysis to include National Basketball Association players. The populations are NHL ($i=1$), NBA ($i=2$),
and EPL ($i=3$). Histograms for the three populations are given in 
Figure \@ref(fig:nhlnbaeplbmi1). The population sizes,
means, and standard deviations are given in Table \@ref(tab:nhlnbaeplbmi1a).

```{r nhlnbaeplbmi1, fig.cap='Distributions of Body Mass Index for NHL, NBA, and EPL players', out.width='80%', fig.asp=.75, fig.align='center', echo=FALSE}

nhl <- read.csv("nhl_ht_wt_1415.csv")
#head(nhl, 2)
nba <- read.csv("nba_ht_wt.csv")
#head(nba, 2)
epl <- read.csv("epl_2015_ht_wt.csv")
#head(epl, 2)

### Compute N, player BMI by league, mean, sd of BMI's by league
N.nhl <- nrow(nhl)
N.nba <- nrow(nba)
N.epl <- nrow(epl)
bmi.nhl <- 703 * nhl$Weight / (nhl$Height^2)
bmi.nba <- 703 * nba$Weight / (nba$Height^2)
bmi.epl <- 703 * epl$Weight / (epl$Height^2)
mu.nhl <- mean(bmi.nhl)
mu.nba <- mean(bmi.nba)
mu.epl <- mean(bmi.epl)
sigma.nhl <- sqrt((N.nhl-1)/N.nhl) * sd(bmi.nhl)
sigma.nba <- sqrt((N.nba-1)/N.nba) * sd(bmi.nba)
sigma.epl <- sqrt((N.epl-1)/N.epl) * sd(bmi.epl)


bmi.grid <- seq(17,33,.01)
par(mfrow=c(3,1))
hist(bmi.nhl, breaks=seq(17,33,.25), freq=F, col="green", xlab="BMI",
     main="NHL BMI")
lines(bmi.grid, dnorm(bmi.grid, mu.nhl, sigma.nhl), col="red", lwd=2)
hist(bmi.nba, breaks=seq(17,33,.25), freq=F, col="blue", xlab="BMI",
     main="NBA BMI")
lines(bmi.grid, dnorm(bmi.grid, mu.nba, sigma.nba), col="red", lwd=2)
hist(bmi.epl, breaks=seq(17,33,.25), freq=F, col="coral", xlab="BMI",
     main="EPL BMI")
lines(bmi.grid, dnorm(bmi.grid, mu.epl, sigma.epl), col="blue", lwd=2)
par(mfrow=c(1,1))
```

```{r nhlnbaeplbmi1a, echo=FALSE, tidy=FALSE}
nhl.nba.epl.out <- rbind(cbind(N.nhl, mu.nhl, sigma.nhl),
                         cbind(N.nba, mu.nba, sigma.nba),
                         cbind(N.epl, mu.epl, sigma.epl))
colnames(nhl.nba.epl.out) <- c("$N$", "$\\mu$", "$\\sigma$")
rownames(nhl.nba.epl.out) <- c("NHL BMI", "NBA BMI", "EPL BMI")
# round(nhl.nba.epl.out, 3)

# summary(c(bmi.nhl, bmi.nba, bmi.epl))

knitr::kable(nhl.nba.epl.out, escape=F, digits=3,
             caption='Population sizes, means, and SDs by league for BMI',
             booktabs=TRUE)

```

$$ N_1=748 \quad N_2=505 \quad N_3=526 \qquad \mu_1=26.514 \quad \mu_2=24.741 \quad \mu_3=23.019 $$
$$\sigma_1=1.449 \quad \sigma_2=1.718 \quad \sigma_3=1.711 $$

While the population standard deviations (and thus variances) are not all equal, a "pooled" (unweighted) variance is used for computational purposes. Also, the 
unweighted mean $\mu$ and  $\alpha_i$ are computed.

$$ \sigma^2 = \frac{1.449^2+1.718^2+1.711^2}{3}=2.660 $$
$$\mu=\frac{26.514+24.741+23.019}{3}=24.758 \qquad
 \alpha_1=26.514-24.758=1.756 $$
 $$ \alpha_2=24.741-24.758=-0.017 \qquad
\alpha_3=23.019-24.758=-1.739 $$

Note that these $\alpha_i$ are obtained under the assumption $\sum \alpha_i=0$. 
If samples of sizes $n_1=n_2=n_3=4$ and $n_1=n_2=n_3=12$ are taken, the following $F$-distributions for the ratio $MST/MSE$ are obtained.

$$ n_i=4: \quad \frac{MST}{MSE} \sim F_{\nu_1,\nu_2,\lambda_1} \quad \lambda_1=
 \frac{4\left(1.756^2+(-0.017)^2+(-1.739)^2\right)}{2.660}=9.185 $$ $$\nu_1=3-1=2 \qquad \qquad \nu_2=12-3=9$$

$$ n_i=12: \quad \frac{MST}{MSE} \sim F_{\nu_1,\nu_2,\lambda_2} \quad \lambda_2=
 \frac{12\left(1.756^2+(-0.017)^2+(-1.739)^2\right)}{2.660}=27.555 $$ 
 $$ \nu_1=3-1=2 \qquad \qquad \nu_2=36-3=33$$


When $n_1=n_2=n_3=4$, the critical value for testing $H_0:\alpha_1=\alpha_2=\alpha_3=0$ at $\alpha=0.05$ significance level is $F_{.05,2,9}=4.256$.
The power of the $F$-test
under this configuration is $\pi_1=.622$.
When $n_1=n_2=n_3=12$, the critical value for testing $H_0:\alpha_1=\alpha_2=\alpha_3=0$ at $\alpha=0.05$ significance level is $F_{.05,2,33}=3.285$.
The power of the $F$-test
under this configuration is $\pi_2=.997$.
The central $F$-density and the non-central $F$-density with sample sizes $n_1=n_2=n_3=4$, $\lambda_1=9.48$  
for the denominator degrees of freedom $\nu_2=9$ is given in 
Figure \@ref(fig:nhlnbaeplbmi2).

```{r nhlnbaeplbmi2, fig.cap='Central (Blue) and non-central (red) $F$-distributions for Body Mass Index example with n=4', out.width='80%', fig.asp=.75, fig.align='center', echo=FALSE}

## Continuation of Previous code chunk nhlnbaeplbmi1

n <- c(4,12)
## Unweighted Mean and Variance
mu.all <- (mu.nhl+mu.nba+mu.epl) / 3
sigma2.all <- (sigma.nhl^2+sigma.nba^2+sigma.epl^2) / 3
alpha.nhl <- mu.nhl-mu.all
alpha.nba <- mu.nba-mu.all
alpha.epl <- mu.epl-mu.all
lambda <- n*(alpha.nhl^2+alpha.nba^2+alpha.epl^2)/sigma2.all
#lambda

df.num <- c(3-1, 3-1)
df.den <- 3*(n-1)
F.crit <- qf(.95, df.num, df.den)
#F.crit

power <- 1-pf(F.crit, df.num, df.den, lambda)
#power

F.grid <- seq(0,30,.01)

plot(F.grid, df(F.grid, df.num[1], df.den[1]), type="l", col="blue", lwd=2,
     xlab="f", ylab="density", main="Case 1: n=4 per League")
lines(F.grid, df(F.grid, df.num[1], df.den[1], lambda[1]), type="l", col="red", lwd=2)
abline(v=F.crit[1], col="purple", lwd=2)
legend("topright", c("Central F", "Non-Central F", "F(.05)"), col=c("blue", "red", "purple"), lwd=2)

```

Now we will take 100000 samples on $n=4$ and $n=12$ players from each league, and conduct the $F$-test 
of equal population means. We will compare the empirical and theoretical power for each case
and compare the distributions of the observed and theoretical $F$-statistics.

```{r nhlnbaeplbmi3, fig.cap='Observed and Theoretical Distributions of the F-statistic for testing equality of mean Body Mass Index for NHL, NBA, and EPL players, with n1=n2=n3=4', out.width='80%', fig.asp=.75, fig.align='center', echo=FALSE}

## Continuation of nhlnbaeplbmi1
set.seed(12345)
n.nhl <- n.nba <- n.epl <- c(4,12)

num.sim <- 100000
ybar.nhl <- matrix(rep(0,2*num.sim), ncol=2)
sd.nhl <- matrix(rep(0,2*num.sim), ncol=2)
ybar.nba <- matrix(rep(0,2*num.sim), ncol=2)
sd.nba <- matrix(rep(0,2*num.sim), ncol=2)
ybar.epl <- matrix(rep(0,2*num.sim), ncol=2)
sd.epl <- matrix(rep(0,2*num.sim), ncol=2)

for(i0 in 1:2) {
  for (i1 in 1:num.sim) {
    y.nhl <- bmi.nhl[sample(N.nhl, n.nhl[i0], replace=FALSE)]
    y.nba <- bmi.nba[sample(N.nba, n.nhl[i0], replace=FALSE)]
    y.epl <- bmi.epl[sample(N.epl, n.epl[i0], replace=FALSE)]
    ybar.nhl[i1,i0] <- mean(y.nhl)
    sd.nhl[i1,i0] <- sd(y.nhl)
    ybar.nba[i1,i0] <- mean(y.nba)
    sd.nba[i1,i0] <- sd(y.nba)
    ybar.epl[i1,i0] <- mean(y.epl)
    sd.epl[i1,i0] <- sd(y.epl)
  }
}

## Print out table of first sample summary stats
first.out <- rbind(cbind(t(ybar.nhl[1,]), t(sd.nhl[1,])),
                   cbind(t(ybar.nba[1,]), t(sd.nba[1,])),
                   cbind(t(ybar.epl[1,]), t(sd.epl[1,])))
colnames(first.out) <- c("ybar(n=4)","ybar(n=12)","sd(n=4)","sd(n=12)")
rownames(first.out) <- c("NHL-sample 1", "NBA-sample 1", "EPL-sample 1")
round(first.out,3)


ybar.all.4 <- (ybar.nhl[,1]+ybar.nba[,1]+ybar.epl[,1])/3
ybar.all.12 <- (ybar.nhl[,2]+ybar.nba[,2]+ybar.epl[,2])/3
SST.4 <- 4*((ybar.nhl[,1]-ybar.all.4)^2 + (ybar.nba[,1]-ybar.all.4)^2 +(ybar.epl[,1]-ybar.all.4)^2)
SSE.4 <- 3*(sd.nhl[,1]^2+sd.nba[,1]^2+sd.epl[,1]^2)
SST.12 <- 12*((ybar.nhl[,2]-ybar.all.12)^2 + (ybar.nba[,2]-ybar.all.12)^2 +(ybar.epl[,2]-ybar.all.12)^2)
SSE.12 <- 11*(sd.nhl[,2]^2+sd.nba[,2]^2+sd.epl[,2]^2)
F.4 <- (SST.4/2) / (SSE.4/9)
F.12 <- (SST.12/2) / (SSE.12/33)
#mean(F.4 >= qf(.95,2,9))
#mean(F.12 >= qf(.95,2,33))

F.grid <- seq(0,30,.01)
hist(F.4[F.4 <= 30], breaks=150, freq=F, col="orange", xlab="f", ylab="density",
     main="F-statistics for n=4, and Non-Central F-distribution")
lines(F.grid, df(F.grid,2,9,9.670), col="blue", lwd=2)
```

```{r nhlnbaeplbmi3a, echo=FALSE, tidy=FALSE}
ybar.s.out <- rbind(cbind(mu.nhl, sigma.nhl, 
                          mean(ybar.nhl[,1]), sd(ybar.nhl[,1]),
                          mean(ybar.nhl[,2]), sd(ybar.nhl[,2])),
                    cbind(mu.nba, sigma.nba, 
                          mean(ybar.nba[,1]), sd(ybar.nba[,1]),
                          mean(ybar.nba[,2]), sd(ybar.nba[,2])),
                    cbind(mu.epl, sigma.epl, 
                          mean(ybar.epl[,1]), sd(ybar.epl[,1]),
                          mean(ybar.epl[,2]), sd(ybar.epl[,2])))
colnames(ybar.s.out) <- c("$\\mu$", "$\\sigma$",  
                          "$\\bar{\\bar{y}}$ $(n=4)$", "$s_{\\bar{y}}$ $(n=4)$", "$\\bar{\\bar{y}}$ $(n=12)$", "$s_{\\bar{y}}$ $(n=12)$")
rownames(ybar.s.out) <- c("NHL", "NBA", "EPL")
#round(ybar.s.out, 3)

knitr::kable(ybar.s.out, escape=F, digits=3,
             caption='Results of sampling, NHL, NBA, EPL BMI',
             booktabs=TRUE)

```

The distributions of the sample means for each league and the two sample sizes are summarized in Table \@ref(tab:nhlnbaeplbmi3a).

Based on 100000 random samples of size $n_i=4$ from each league, the $F$-test rejected the null hypothesis of no league differences in 63.98\%
of the samples. With samples of size $n_i=12$, 99.45\% of the $F$-tests rejected the null hypothesis. Despite the fact that the
populations of measurements are not exactly normally distributed with equal variances, the test performs as expected, see 
Figure \@ref(fig:nhlnbaeplbmi3).
Computations for the first samples of size $n_1=n_2=n_3=12$ are given below.

$$ \overline{y}_{1.}=25.938 \quad \overline{y}_{2.}=24.635 \quad \overline{y}_{3.}=22.940 \quad \overline{y}_{..}=24.504 \qquad
s_1=1.972 \quad s_2=2.446 \quad s_3=2.068 $$

$$ SST=12\left[(25.938-24.504)^2 + (24.635-24.504)^2 + (22.940-24.504)^2\right] = 54.235 $$
$$  df_T=3-1=2 \qquad MST=\frac{54.235}{2}=27.118 $$

$$ SSE = (12-1)\left[1.972^2 + 2.446^2 + 2.068^2 \right] = 155.632 $$ 
$$ df_E=3(12)-3=33 \quad MSE=\frac{155.632}{33}=4.716 $$

$$ H_0:\mu_1=\mu_2=\mu_3 \qquad \mbox{TS: } F_{obs}=\frac{27.118}{4.716}=5.75 $$ 
$$\mbox{RR: } F_{obs} \geq 3.285 \quad P=P\left(F_{2,33} \geq 5.75\right) =.0072 $$

$$ \nabla $$


**Example 7.2: Comparison of 5 Mosquito Repellents**

A study compared $k=5$ mosquito repellent patches on fabric for soldiers in military operations [@bhatnagar_2007].
The 5 treatments were: Odomos (1), Deltamethrin (2), Cyfluthrin (3), Deltamethrin+Odomos (4), and Cyfluthrin+Odomos (5), with $n_i=30$
subjects per treatment, and a total of $N=150$ measurements. The response observed was the ``Per Man-Hour Mosquito Catch.''
Sample statistics are given in Table \@ref(tab:mosquito1), and the Analysis of Variance is given in Table \@ref(tab:mosquito2). 
Data that have been generated to match the means and standard deviations are plotted in Figure \@ref(fig:mosquito3). The overall mean (long line)
and individual treatment means (short lines) are included.



```{r mosquito1, echo=FALSE, tidy=FALSE}

trt.name <- c("Odomos (1)", "Deltamethrin (2)", "Cyfluthrin (3)", "D+O (4)", "C+O (5)")
n <- rep(30,5)
ybar <- c(7.900, 8.133, 8.033, 6.333, 5.367)
sd <- c(3.367, 3.461, 3.011, 3.122, 3.068)
mosquito1 <- data.frame(trt.name, n, ybar, sd)
colnames(mosquito1) <- c("Treatment ($i$)", "$n$", "$\\bar{y}_i$", "$s_i$")

knitr::kable(mosquito1, escape=F,
             caption='Summary statistics for mosquito repellant experiment',
             booktabs=TRUE)

```
```{r mosquito2, echo=FALSE, tidy=FALSE}

source <- c("Treatments", "Error", "Total")
df <- c(4,145,149)
ss <- c(184.560,1494.680,1679.334)
ms <- c(46.163,10.308,"   ")
f_obs <- c(4.478,"   ","   ")
f.05 <- c(round(qf(.95,4,145),3),"   ","   ")
f.p <- c(round(1-pf(4.478,4,145),3),"   ","   ")
mosquito2 <- data.frame(source,df,ss,ms,f_obs,f.05,f.p)
colnames(mosquito2) <- c("Source", "df", "SS", "MS", "$F_{obs}$", "$F_{.05,4,145}$", "$P(F_{4,145}>F_{obs})$")

knitr::kable(mosquito2, escape=F, digits=3,
            caption='The Analysis of Variance table for the Mosquito Repellent study',
            booktabs=TRUE)

```


```{r mosquito3, fig.cap='Mosquito catch by repellent treatment - data generated to match treatment means and standard deviations', out.width='80%', fig.asp=.75, fig.align='center', echo=FALSE}

mp <- read.csv("mosquito_patch.csv")
head(mp)

plot(mp$y.mosq ~ mp$trt.mosq, main="Mosquito Contact by Repellant",
   pch=16, xaxt="n", col="purple", cex=0.75, xlim=c(0.5,5.5),
   xlab="Repellant", ylab="Per man-hour Mosquito catch")
axis(1,1:5,labels=c("O", "D", "C", "D+O", "C+O"))
lines(c(0.5, 5.5), c(mean(mp$y.mosq),mean(mp$y.mosq)), col="green2", lwd=2)
for (i in 1:length(unique(mp$trt.mosq))) {
  lines(c(i-0.25,i+0.25), 
   c(mean(mp$y.mosq[mp$trt.mosq==i]),mean(mp$y.mosq[mp$trt.mosq==i])),
    col="blue", lwd=2)
}

mod1 <- aov(y.mosq ~ factor(trt.mosq), data=mp)
anova(mod1)
```

 

- $H_0: \alpha_1 = \alpha_2 = \alpha_3 = \alpha_4=\alpha_5= 0  \quad (\mu_1 = \mu_2 = \mu_3=\mu_4=\mu_5)$
 (No treatment effect)
- $H_A: \mbox{ Not all } \alpha_i$ are 0 (Treatment effects exist)
- T.S. $F_{obs} = \frac{MST}{MSE}=4.478$
- R.R.: $F_{obs} > F_{\alpha, k-1, n-k}=F_{0.05,4,145}=2.434$
- p-value: $P=P(F_{k-1,N-k} \geq F_{obs})=P(F_{4,145}\geq 4.478)=.0019$


Since the $F$-statistic is sufficiently large, conclude that the means differ,
then the following methods can used to make comparisons among treatments.

$$ \nabla $$ 

### Comparisons among Treatment Means
 
Assuming that it has been concluded that treatment means differ, we
generally would like to know which means are significantly different.
This is generally done by making contrasts among treatments.
Special cases of contrasts include pre-planned or all pairwise
comparisons between pairs of treatments. 

A **contrast** is a linear function of treatment means, where the coefficients sum to 0. 
A contrast among population means can be estimated with the same contrast among sample means, and inferences can be made based on the sampling
distribution of the contrast. Let $C$ be the contrast among the population means, and $\hat{C}$ be its estimator based on means of the independent
random samples.

$$ C= a_1\mu_1 + \cdots + a_k\mu_k = \sum_{i=1}^k a_i \mu_i \mbox{  where  }  \sum_{i=1}^k a_i = 0 \qquad
\hat{C} = a_1\overline{Y}_{1.} + \cdots + a_k\overline{Y}_{k.} = \sum_{i=1}^k a_i \overline{Y}_{i.}  $$ 
$$ V\{\hat{C}\} = \sigma^2\left[\frac{a_1^2}{n_1} + \cdots + \frac{a_k^2}{n_k}\right] = \sigma^2 \sum_{i=1}^k \frac{a_i^2}{n_i} $$

When the sample sizes are balanced (all $n_i$ are equal), the formula for the variance clearly simplifies.
Contrasts are specific to particular research questions, so the general rules for tests and Confidence Intervals are given here, followed
by an application to the Mosquito Repellent study. Since the coefficients sum to 0, we are virtually always testing $H_0: C=0$ in practice.

Once samples are obtained, obtain $\hat{c}$, the contrast based on the observed sample means among the treatments.  

$$ \hat{c} = a_1\overline{y}_{1.}+ \cdots + a_k\overline{y}_{k.}=\sum_{i=1}^k a_i\overline{y}_{i.}  \qquad
\widehat{SE}\{\hat{C}\}=\sqrt{MSE \sum_{i=1}^k \frac{a_i^2}{n_i}} $$

Testing whether a contrast is equal to 0 and obtaining a $(1-\alpha)100\%$ Confidence Interval for $C$ are done as follow.

$$ H_0: C=0 \quad H_A: C\neq 0 \quad \mbox{TS: } t_C = \frac{\hat{c}}{\widehat{SE}\{\hat{C}\}}  \quad \mbox{RR: } |t_C| \geq t_{\alpha/2,N-k} \quad
 P=2P\left(t_{N-k} \geq |t_C|\right)  $$

$$ (1-\alpha)100\% \mbox{ Confidence Interval for } C: \qquad \hat{c} \pm t_{\alpha/2,N-k} \widehat{SE}\{\hat{C}\} $$

The test can be conducted as upper or lower-tailed with obvious adjustments.
An alternative approach is to compute the sums of squares for the contrast $SSC$, and use an $F$-test, comparing its Mean Square to $MSE$.

$$ SSC=\frac{\left(\hat{c}\right)^2}{\sum_{i=1}^k \frac{a_i^2}{n_i}}  \qquad  df_C=1 \qquad  MSC=\frac{SSC}{1}=SSC  $$


$$ H_0: C=0 \qquad H_A: C\neq 0 \qquad \mbox{TS: } F_C=\frac{MSC}{MSE} $$ $$ \mbox{RR: } F_C \geq F_{\alpha,1,N-k}  \qquad \qquad
 P=P\left(F_{1,N-k} \geq F_C \right) $$ 

**Example 7.3: Comparison of 5 Mosquito Repellents - Contrasts 1 and 2**

Suppose the researchers are interested in comparing the two treatments that use Deltamethrin (2 and 4) with the two treatments
that use Cyfluthrin (3 and 5). Then, the following calculations are made.

$$ C_1 = \left(\mu_2+\mu_4\right) - \left(\mu_3+\mu_5\right) \qquad a_1=0 \quad a_2=a_4=1 \quad a_3=a_5=-1 \qquad n_{i}=30 \qquad MSE=10.308$$

$$ \overline{y}_{2.}=8.133 \quad \overline{y}_{4.}=6.333  \quad \overline{y}_{3.}= 8.033 \quad \overline{y}_{5.}= 5.367 \qquad
\hat{c}_1 = (8.133+6.333)-(8.033+5.367)=1.066  $$

$$ \widehat{SE}\{\hat{C}_1\} = \sqrt{10.308\left(\frac{0^2+1^2+(-1)^2+1^2+(-1)^2}{30}\right)}=1.172 $$

For a test ($\alpha=0.05$) of $H_0:C_1=0$, the test statistic, rejection region and p-value, along with a 95\% Confidence Interval for 
$C$ are given below.

$$ \mbox{TS: } t_{C_1} = \frac{1.066}{1.172} = 0.910 \quad \mbox{RR: } |t_{C_1}| \geq t_{.025,145}=1.976 \quad P=2P\left(t_{145} \geq |0.910|\right)=.364 $$

$$ \mbox{95}\% \mbox{ CI for } C: 1.066 \pm 1.976(1.172) \equiv 1.066 \pm 2.316 \equiv (-1.250,3.382) $$

There is no evidence of any difference between the effects of these two types of repellents. Next, we conduct the $F$-test, knowing in advance that its
conclusion and p-value will be equivalent to 2-tailed $t$-test performed above (the only difference due to rounding is in third decimal place).

$$ SSC_1=\frac{1.066^2}{\frac{4}{30}} = 8.523 = MSC_1  \quad \mbox{TS: } F_{C_1}=\frac{8.523}{10.308}=0.827 $$
$$\mbox{RR: } F_{C_1}\geq F_{.05,1,145}=3.906 \qquad
P = P\left(F_{1,145} \geq 0.827 \right) = .3647$$
$$ P=P\left(F_{1,145}\geq 0.827\right)=.365 $$

For a second contrast $(C_2)$, without going through all calculations, consider comparing Deltamethrin and Cyfluthrin (each without Odomos: 2 and 3)
with Deltamethrin and Cyfluthrin (each with Odomos: 4 and 5). This involves: $a_1=0,a_2=a_3=1,a_4=a_5=-1$. Note that the standard error of the
contrast will be exactly the same as that for contrast $\hat{c}_1$. 

$$ \hat{c}_2 = 4.466 \quad TS:t_{C_2}=\frac{4.466}{1.172}=3.811 \quad P=2P\left(t_{145}\geq|3.811|\right)=.0002 \qquad
\mbox{95}\% \mbox{ CI: } 4.466\pm 2.316 \equiv(2.150,6.782)   $$

There is evidence that the combined mean is higher without Odomos than with Odomos. Since low values are better (mosquito contacts),
Odomos as an additive to the two chemicals (individually) is better than no additive to the two chemicals individually. 
The $F$-test is given below. The R code that follows extends the calculations made in Example 7.2.

$$ SSC_2=\frac{4.466^2}{\frac{4}{30}} = 149.59   \quad 
\mbox{TS: } F_{C_2}=\frac{149.59}{10.308}=14.51 
\quad P=P\left(F_{1,145}\geq 14.51\right)=.0005 $$

```{r mosquito4, echo=FALSE}
## Continuation of mosquito3

n.trt <- as.numeric(tapply(mp$y.mosq, mp$trt.mosq, length))    ## sample size by trt
ybar.trt <- as.numeric(tapply(mp$y.mosq, mp$trt.mosq, mean))   ## sample mean by trt
sd.trt <- as.numeric(tapply(mp$y.mosq, mp$trt.mosq, sd))       ## sample sd by trt

SSE <- sum((n.trt-1)*sd.trt^2)      ## Error Sum of Squares
dfE <- sum(n.trt) - length(n.trt)   ## Error degrees of freedom
MSE <- SSE / dfE                    ## Error Mean Square

a1 <- c(0,1,-1,1,-1)       ## Contrast 1: (D,DO) - (C,CO)
a2 <- c(0,1,1,-1,-1)       ## Contrast 2: (D,C) - (DO,CO)

c1.hat <- sum(a1*ybar.trt)                       ## c1-hat
SE.c1.hat <- sqrt(MSE * sum(a1^2/n.trt))         ## SE{c1-hat}
t.c1 <- c1.hat / SE.c1.hat                       ## t-stat for H0: C1=0
SSC1 <- c1.hat^2 / sum(a1^2/n.trt)               ## C1 sum of Squares
F.c1 <- (SSC1/1) / MSE                           ## F-stat for H0: C1=0

c2.hat <- sum(a2*ybar.trt)                       ## c2-hat
SE.c2.hat <- sqrt(MSE * sum(a2^2/n.trt))         ## SE{c2-hat}
t.c2 <- c2.hat / SE.c2.hat                       ## t-stat for H0: C2=0
SSC2 <- c2.hat^2 / sum(a2^2/n.trt)               ## C2 sum of Squares
F.c2 <- (SSC2/1) / MSE                           ## F-stat for H0: C2=0

## Output Results of tests, CIs

contrast12.out <- rbind(
                  cbind(c1.hat,SE.c1.hat, t.c1, 2*(1-pt(abs(t.c1),dfE)),SSC1,F.c1, 1-pf(F.c1,1,dfE),
                        c1.hat-qt(.975,dfE)*SE.c1.hat, c1.hat+qt(.975,dfE)*SE.c1.hat),
                  cbind(c2.hat,SE.c2.hat, t.c2, 2*(1-pt(abs(t.c2),dfE)),SSC2,F.c2, 1-pf(F.c2,1,dfE),
                        c2.hat-qt(.975,dfE)*SE.c2.hat, c2.hat+qt(.975,dfE)*SE.c2.hat)
)
colnames(contrast12.out) <- c("Estimate", "Std Err", "t_obs", "P(>|t_{obs}|)",
                            "SS", "F_obs", "P(>F_obs)", "LB", "UB")
rownames(contrast12.out) <- c("C1: (D,DO)-(C,CO)", "C2: (D,C)-(DO,CO)")
round(contrast12.out,3)

```

$$   \nabla   $$

A special class of contrasts are **orthogonal contrasts**. Two contrasts are orthogonal if the sum of the products of their $a_i$ coefficients,
divided by the sample sizes $n_i$, is 0.
This concept is shown below.

$$ C_1 =\sum_{i=1}^k a_{1i}\mu_i \qquad C_2 =\sum_{i=1}^k a_{2i}\mu_i  \quad \mbox{$C_1$ and $C_2$ are orthogonal if } \quad
\sum_{i=1}^k \frac{a_{1i}a_{2i}}{n_i}=0 $$

Note that if the sample sizes are all equal (balanced design), this simplifies to $\sum_{i=1}^ka_{1i}a_{2i}=0$. The two contrasts in Example 7.3 are orthogonal
(check this). If there are $k$ treatments, and $k-1$ degrees of freedom for Treatments, any $k-1$ pairwise orthogonal contrasts' sums of squares
will add up to the Treatment sum of squares. That is, $SST$ can be decomposed into the sums of squares for the $k-1$ contrasts. The decomposition is not unique,
there may be various sets of orthogonal contrasts.

**Example 7.4: Comparison of 5 Mosquito Repellents - Orthogonal Contrasts**

Consider these two other contrasts.

- (D versus C without O) vs (D versus C with Odomos): $C_3=\left(\mu_2-\mu_3\right)-\left(\mu_4-\mu_5\right) = \mu_2-\mu_3-\mu_4+\mu_5$
- Odomos only versus the four other treatments: $C_4=4\mu_1 - \mu_2 - \mu_3 - \mu_4 - \mu_5$


The following R code chunk gives the contrast coefficients for these four contrasts. For all six pairs, $\sum_{i=1}^k a_{ji} a_{j'i} =0$, $j\neq j'$.
Also given are the estimates, standard errors, $t$-tests, 95\% Confidence Intervals, Sums of Squares and $F$-statistics.

```{r mosquito5, echo=FALSE}
## Continuation of mosquito4
a3 <- c(0,1,-1,-1,1)       ## Contrast 3: (D-C) - (DO-CO)
a4 <- c(4,-1,-1,-1,-1)       ## Contrast 4: 4O - (D,C,DO,CO)

a11 <- c(a1, sum(a1))
a22 <- c(a2, sum(a2))
a33 <- c(a3, sum(a3))
a44 <- c(a4, sum(a4))
a12 <- c(a1*a2, sum(a1*a2))
a13 <- c(a1*a3, sum(a1*a3))
a14 <- c(a1*a4, sum(a1*a4))
a23 <- c(a2*a3, sum(a2*a3))
a24 <- c(a2*a4, sum(a2*a4))
a34 <- c(a3*a4, sum(a3*a4))
a.df <- data.frame(a11,a22,a33,a44,a12,a13,a14,a23,a24,a34)
colnames(a.df) <- c("a1", "a2", "a3", "a4", "a1a2", "a1a3", "a1a4", "a2a3", "a2a4", "a3a4")
rownames(a.df) <- c("O", "D", "C", "DO", "CO", "sum")
a.df

c3.hat <- sum(a3*ybar.trt)                       ## c3-hat
SE.c3.hat <- sqrt(MSE * sum(a3^2/n.trt))         ## SE{c3-hat}
t.c3 <- c3.hat / SE.c3.hat                       ## t-stat for H0: C3=0
SSC3 <- c3.hat^2 / sum(a3^2/n.trt)               ## C3 sum of Squares
F.c3 <- (SSC3/1) / MSE                           ## F-stat for H0: C3=0

c4.hat <- sum(a4*ybar.trt)                       ## c4-hat
SE.c4.hat <- sqrt(MSE * sum(a4^2/n.trt))         ## SE{c4-hat}
t.c4 <- c4.hat / SE.c4.hat                       ## t-stat for H0: C4=0
SSC4 <- c4.hat^2 / sum(a4^2/n.trt)               ## C4 sum of Squares
F.c4 <- (SSC4/1) / MSE                           ## F-stat for H0: C4=0

## Output Results of tests, CIs

contrast34.out <- rbind(
                  cbind(c3.hat,SE.c3.hat, t.c3, 2*(1-pt(abs(t.c3),dfE)),SSC3, F.c3, 1-pf(F.c3,1,dfE),
                        c3.hat-qt(.975,dfE)*SE.c3.hat, c3.hat+qt(.975,dfE)*SE.c3.hat),
                  cbind(c4.hat,SE.c4.hat, t.c4, 2*(1-pt(abs(t.c4),dfE)),SSC4, F.c4, 1-pf(F.c4,1,dfE),
                        c4.hat-qt(.975,dfE)*SE.c4.hat, c4.hat+qt(.975,dfE)*SE.c4.hat)
)
#colnames(contrast34.out) <- c("Estimate", "Std Err", "t_obs", "P(>|t_{obs}|)",
          #                  "F_obs", "P(>F_obs)", "LB", "UB")
rownames(contrast34.out) <- c("C3: (D-C) - (DO-CO)", "C4: 4O-(D,C,DO,CO)")
contrast.out <- rbind(contrast12.out, contrast34.out)
round(contrast.out,3)
```


From the original Analysis of Variance table, see that the Treatment sum of squares is $SST=184.65$. As these four contrasts are pairwise orthogonal,
their sums of squares add up to $SST$: $8.533+149.544+5.625+20.944=184.646$. Note that virtually all of the differences among the treatments
(based on this set of contrasts) is contrast 2, comparing the average of D and C without O versus the average of D and C with O.


$$ \nabla  $$



#### Bonferroni Method of Multiple Comparisons

The Bonferroni method is used in many situations and is based on the following
premise: If there are $c$ comparisons or contrasts to be made simultaneously, and we desire to be $\left(1-\alpha_E\right)100\%$
confident that all are correct,  each comparison should be made at a
higher level of confidence (lower probability of type I error). If 
individual comparisons are made at $\alpha_I=\alpha_E/c$ level of significance, there is an
overall error rate no larger than $\alpha_E$. This method is conservative
and can run into difficulties (low power) as the number of comparisons
gets very large. The general procedures for simultaneous tests and Confidence Intervals are as follow in terms of comparing various contrasts.

$$ \mbox{Define: } \quad B_{C_j} = t_{\alpha_E/(2c),\nu} \widehat{SE}\left\{\hat{C_j}\right\}  \quad j=1,\ldots,c$$

$$ \mbox{Conclude } C_j\neq 0 \mbox{ if } \left|\hat{c_j}\right| \geq B_{C_j} $$
$$ \mbox{Simultaneous }\left(1-\alpha_E\right)100\% \mbox{ CI's for } C_j: \quad  (\hat{c_j}) \pm  B_{C_j} \quad j=1,\ldots,c$$

where $t_{\alpha_E /(2c),\nu}$, with $\nu$ being the error degrees of freedom, $\nu=N-k$ for the Completely Randomized Design, 
is obtained from the Bonferroni $t$--table (see chapter powerpoint slides) or from statistical packages or EXCEL. 

The Bonferroni adjusted p-value can be computed as the minimum of 1
and $c$ times the unadjusted p-value.

**Example 7.5: Comparison of 5 Mosquito Repellents - Bonferroni CIs for Contrasts**

In this example, we will obtain Bonferroni simultaneous Confidence Intervals for the $c=4$ contrasts considered in Example 7.4. The
adjustment involves replacing $t_{.05/2,N-k}=1.976$ with $t_{.05/(2(4)),N-k}=2.529$. Recall that the estimated standard
errors for contrasts 1, 2, and 3 are all equal to  1.172 and the standard error for contrast 4 is 2.621. This leads to the following values for
$B_{C_j} \quad j=1,2,3,4$.

$$ B_{C_1}=B_{C_2}=B_{C_3}=2.529(1.172)=2.964 \qquad
B_{C_4}=2.529(2.621)=6.629$$

The following R code chunk computes the original 95\% CI's for the contrasts as well as the Bonferroni adjusted 95\% CI's. 

```{r mosquito6, echo=FALSE}
## Continuation of mosquito5
c <- 4                                ## Number of contrasts
t.025 <- qt(1-.05/2,dfE)              ## t_(.05/2,dfE)
Bon.t <- qt(1-.05/(2*c),dfE)          ## t_(.05/(2c), dfE)

## Obtain unadjusted P_values for t-tests for Contrasts
p.c1 <- 2*(1-pt(abs(t.c1),dfE))  
p.c2 <- 2*(1-pt(abs(t.c2),dfE)) 
p.c3 <- 2*(1-pt(abs(t.c3),dfE)) 
p.c4 <- 2*(1-pt(abs(t.c4),dfE)) 
   
## SEt up and print table of output        
bon.contrast.out <- rbind(
  cbind(c1.hat, SE.c1.hat, t.c1, p.c1, min(c*p.c1,1),
        c1.hat-t.025*SE.c1.hat, c1.hat+t.025*SE.c1.hat, 
        c1.hat-Bon.t*SE.c1.hat, c1.hat+Bon.t*SE.c1.hat),
  cbind(c2.hat, SE.c2.hat, t.c2, p.c2, min(c*p.c2,1),
        c2.hat-t.025*SE.c2.hat, c2.hat+t.025*SE.c2.hat, 
        c2.hat-Bon.t*SE.c2.hat, c2.hat+Bon.t*SE.c2.hat),
  cbind(c3.hat, SE.c3.hat, t.c3, p.c3, min(c*p.c3,1),
        c3.hat-t.025*SE.c3.hat, c3.hat+t.025*SE.c3.hat, 
        c3.hat-Bon.t*SE.c3.hat, c3.hat+Bon.t*SE.c3.hat),
  cbind(c4.hat, SE.c4.hat, t.c4, p.c4, min(c*p.c4,1),
        c4.hat-t.025*SE.c4.hat, c4.hat+t.025*SE.c4.hat, 
        c4.hat-Bon.t*SE.c4.hat, c4.hat+Bon.t*SE.c4.hat)
)

colnames(bon.contrast.out) <- c("Estimate", "Std Err", "t_obs", "P(>|t_{obs}|)", "Bon Adj P", "LB", "UB", "Bon LB", "Bon UB")
rownames(bon.contrast.out) <- c("C1: (D,DO)-(C,CO)", "C2: (D,C)-(DO,CO)",
  "C3: (D-C) - (DO-CO)", "C4: 4O-(D,C,DO,CO)")
round(bon.contrast.out, 3)
```
Note that the Bonferroni simultaneous Confidence Intervals are wider than the unadjuted ones, and the adjusted p-values are higher, but Contrast 2 is still significant.

$$ \nabla $$

#### Tukey Method for All Pairwise Comparisons
 
Various methods have been
developed to handle all possible pairwise comparisons 
among treatment means
and keep the overall error
rate at $\alpha_E$, including the widely reported Bonferroni procedure
described above. 
Note that pairwise comparisons are a special class of contrasts where
one $\alpha_i=1$, one $\alpha_i=-1$ and the remaining $k-2 \hspace{0.1in} \alpha_i=0$.
When there are $k$ treatments, there are $k(k-1)/2$ pairs of treatments that can be compared (combinations from Chapter 3).

One commonly used procedure is **Tukey's Honest Significant Difference (HSD)** method,
which is more powerful than the Bonferroni method (but more limited
in its applicability).
Statistical computer packages can make 
these comparisons automatically.
Tukey's method can be used for tests and confidence intervals for all pairs of
treatment means simultaneously. If there are $k$ treatments, their will
be $\frac{k(k-1)}{2}$ such tests or intervals. The general forms, allowing for
different sample sizes for treatments $i$ and $i'$ are as follow (the unequal sample size case is referred to as the "Tukey-Kramer" method).
The procedure makes use of the **Studentized Range Distribution** with critical values, $q_{\alpha_E,k,\nu}$, indexed by the number of treatments $(k)$ and
error degrees of freedom $\nu = N-k$ for the Completely Randomized Design. 
The R functions **qtukey** and **ptukey** give quantiles and probabilities for the
distribution. A table of critical values for $\alpha_E=0.05$ is given in this chapter's powerpoint slides.

$$ \mbox{Define: } \quad HSD_{ii'} = \frac{q_{\alpha_E,k, \nu}}{2} \widehat{SE}\{\overline{Y}_{i.} - \overline{Y}_{i'.}\} = \frac{q_{\alpha_E,k, \nu}}{2} \sqrt{MSE
\left(\frac{1}{n_i} +
\frac{1}{n_{i'}}\right)} \quad i\neq i' $$

$$ \mbox{Conclude } \mu_i \neq \mu_{i'} \quad \mbox{ if } \quad  \left|\overline{y}_{i.}-\overline{y}_{i'.}\right| \geq HSD_{ii'} $$
$$  \mbox{Simultaneous }\left(1-\alpha_E\right)100\% \mbox{ CI's for } \mu_i-\mu_{i'}: \quad  \left(\overline{y}_{i.} - \overline{y}_{i'.}\right) \pm  HSD_{ii'} $$

 
When the sample sizes are equal ($n_i=n_{i'}$), the formula for $HSD_{ii'}$
can be simplified as follows.

$$ HSD_{ii'} = q_{\alpha_E,k, \nu} \sqrt{MSE
\left(\frac{1}{n_i}\right)} \quad i<i'
$$ 


**Example 7.6: Comparison of 5 Mosquito Repellents**

The Bonferroni and
Tukey methods are used to obtain simultaneous 95\% CI's for each difference in mean mosquito contacts. The general form for the Bonferroni
simultaneous 95\% CI's (with $c=5(4)/2=10$ and $\nu=150-5=145)$) is given below. Recall that $MSE=10.308$ and $n_i=30$ for each treatment.

$$ B_{ii'} = t_{.05/(2(10)),145} \sqrt{10.308
\left(\frac{1}{30} +
\frac{1}{30}\right)} = 2.851(0.829)=2.363 $$
$$\mbox{Simultaneous }95\% \mbox{ CIs: } \quad \left(\overline{y}_{i.} - \overline{y}_{i'.}\right) \pm 2.363$$


For Tukey's method, the confidence intervals are of the following form. 

$$ HSD_{ii'}=q_{0.05, 5, 145}\sqrt{10.308\left(\frac{1}{30}\right)} = 3.907(0.586) = 2.290 $$
$$ \qquad \mbox{Simultaneous } 95\% \mbox{ CIs: } \quad\left(\overline{y}_{i.} - \overline{y}_{i'.}\right) \pm 2.290$$


The corresponding calculations and confidence intervals are given in in the following R code chunk.

```{r mosquito7, echo=FALSE}
## Continuation of mosquito6

k <- length(n.trt)                ## number of treatments
c.pair <- k*(k-1)/2               ## number of pairwise comparisons
q.05 <- qtukey(1-.05,k,dfE)       ## Critical Value for Studentized Range
Bon.t <- qt(1-.05/(2*c.pair),dfE) ## Critical value for Bonferroni

HSD <- q.05*sqrt(MSE/n.trt[1])         ## Tukey HSD (Balanced n)
Bon.B <- Bon.t*sqrt(2*MSE/n.trt[1])    ## Bonferroni B (Balanced n)

## Create a matrix to hold results for Simultaneous CI's and row counter
sim.ci.out <- matrix(rep(0,c.pair*11), ncol=11)
sim.row <- 0

## Loop through all pairs (higher label trt is "i", lower is "i")
for (i2 in 2:k) {
  for (i1 in 1:(i2-1)) {
    sim.row <- sim.row + 1   ## Row of table
    sim.ci.out[sim.row,1] <- i2    ## Trt i
    sim.ci.out[sim.row,2] <- i1    ## Trt i'
    sim.ci.out[sim.row,3] <- ybar.trt[i2]  ## Trt i mean
    sim.ci.out[sim.row,4] <- ybar.trt[i1]  ## Trt i' mean
    sim.ci.out[sim.row,5] <- ybar.trt[i2]-ybar.trt[i1]  ## mean diff
    sim.ci.out[sim.row,6] <- HSD           ## Tukey HSD
    sim.ci.out[sim.row,7] <- sim.ci.out[sim.row,5] - HSD ## Tukey LB
    sim.ci.out[sim.row,8] <- sim.ci.out[sim.row,5] + HSD ## Tukey UB
    sim.ci.out[sim.row,9] <- Bon.B        ## Bonf B
    sim.ci.out[sim.row,10] <- sim.ci.out[sim.row,5] - Bon.B ## Bon LB
    sim.ci.out[sim.row,11] <- sim.ci.out[sim.row,5] + Bon.B ## Bon UB
  }
}
colnames(sim.ci.out) <- c("i", "i'", "yb i", "yb i'", "diff",
                          "HSD", "T LB", "T UB", "Bon B", "B LB",
                          "B UB")
round(sim.ci.out,3)

## Tukey HSD using built-in R function
mosq.mod1 <- aov(y.mosq ~ factor(trt.mosq), data=mp)
anova(mosq.mod1)
TukeyHSD(mosq.mod1)

```

Based on the intervals in the previous R code chunk, it can be concluded that
treatments 1 (Odomos) and 5 (Cyfluthrin + Odomos) are significantly different, as are treatments 2 (Deltamethrin) and 5,
and treatments 3 (Cyfluthrin) and 5. 

While it is easy to write a function in R to conduct the Bonferroni method, there does not appear an easy "follow up" to the ANOVA
that prints out actual Confidence Intervals (in base R,
anyway). There 
is an easy one for Tukey's honest significant difference method, the **TukeyHSD** function (see above). Note that R takes the mean
with the higher subscript minus the mean with the lower subscript.

$$ \nabla $$



## Randomized Block Design (RBD) For Studies Based on Matched Units
 
In crossover designs (aka within subjects designs), each unit or subject receives each treatment. In
these cases, units are referred to as **blocks**. In other studies, units or subjects may be matched based on external criteria.
The notation for the Randomized Block Design (RBD) is very similar to that of the
CRD, with additional elements.
The model we are assuming here is written as follows.

$$ Y_{ij} = \mu + \alpha_i + \beta_j + \epsilon_{ij} = \mu_i +\beta_j
+\epsilon_{ij} \quad i=1,\ldots,k; \quad j=1,\ldots,b  $$

Here, $\mu$ represents the overall mean measurement, $\alpha_i$ is
the (**fixed**) effect of the $i^{th}$ treatment, $\beta_j$ is the 
(typically **random**) effect of the
$j^{th}$ block, and $\epsilon_{ij}$ is a random error component 
that can be thought of as the 
variation in measurements if the same experimental unit received the same
treatment repeatedly. Note that just as before, $\mu_i$ represents the
mean measurement for the $i^{th}$ treatment (across all blocks). The
general situation will consist of an experiment with $k$ treatments
being received by each of $b$ blocks. Blocks can be fixed or random, typically they are random.


\subsection{Test Based on Normally Distributed Data}

When the (random) block effects ($\beta_j$) and random error terms
($\epsilon_{ij}$) are independent and 
normally distributed, an $F$-test is conducted that is
similar to that described for the Completely Randomized Design, but with an extra source of variation.
If blocks are fixed, the analysis is the same.
The notation used is as
follows. 


$$ N = bk \qquad
\overline{y}_{i.} = \frac{\sum_{j=1}^{b}y_{ij}}{b} \qquad
\overline{y}_{.j} = \frac{\sum_{i=1}^{k}y_{ij}}{k} \qquad
\overline{y}_{..} = \frac{\sum_{i=1}^k \sum_{j=1}^{b}y_{ij}}{bk} $$
$$ \mbox{Total SS: } \quad TSS =\sum_{i=1}^k \sum_{j=1}^{b} \left(y_{ij} - \overline{y}_{..} \right)^2 \qquad df_{\mbox{Tot}}=N-1=bk-1$$
$$ \mbox{Treatment SS:} \quad SST =b\sum_{i=1}^k  \left(\overline{y}_{i.} - \overline{y}_{..} \right)^2 \qquad df_T=k-1$$ 
$$ \mbox{Block SS:} \quad SSB =k\sum_{j=1}^b  \left(\overline{y}_{.j} - \overline{y}_{..} \right)^2 \qquad df_B=b-1$$
$$ \mbox{Error SS: } \quad
SSE = \sum_{i=1}^k \sum_{j=1}^{b} \left(y_{ij} - \overline{y}_{i.} - \overline{y}_{.j}  + \overline{y}_{..} \right)^2  \qquad df_E
= (b-1)(k-1) $$


Note that the Analysis of Variance simply has added items representing the block means
$\left(\overline{y}_{.j}\right)$ and variation among the block means ($SSB$). We can further
think of this as decomposing the total variation into differences among the
treatment means ($SST$), differences among the block means ($SSB$), and
random variation not explained by either differences among treatment or block means ($SSE$). Also, note that $SSE=TSS-SST-SSB$.


```{r rbdaov1, echo=FALSE, tidy=FALSE}
aovtable.out <- rbind(
cbind("Treatments", "k-1", "SST", "MST=SST/(k-1)", "F=MST/MSE"),
cbind("Blocks", "b-1", "SSB", "MSB=SSB/(b-1)", "F=MSB/MSE"),
cbind("Error", "(b-1)(k-1)", "SSE", "MSE=SSE/[(b-1)(k-1)]", "   "),
cbind("Total", "N-1", "TSS", "   ","   ")
)
colnames(aovtable.out) <- c("Source", "Degrees of Freedom", "Sum of Squares", "Mean Square", "$F_{obs}$")

knitr::kable(aovtable.out, escape=FALSE, caption='The Analysis of Variance Table for the Randomized Block Design',
             booktabs=TRUE)

```

 
Once again, the main purpose for conducting this type of experiment is
to detect differences among the treatment means (treatment effects). The
test is very similar to that of the CRD, with only minor adjustments. We
often are not interested in testing for differences among blocks, since we
expect there to be differences among them (that's why the
design was set up this way), and they were just a random
sample from a population of such experimental units. However, in some cases, estimating the unit to unit (subject to subject) variance component is of interest.
The testing procedure can be described as follows.

- $H_0: \alpha _1 = \cdots = \alpha_k = 0  \quad (\mu_1 = \cdots = \mu_k)$
 (No treatment effect)
- $H_A: \mbox{ Not all } \alpha_i$ are 0 (Treatment effects exist)
- T.S. $F_{obs} = \frac{MST}{MSE}$
- R.R.: $F_{obs} \geq F_{\alpha, k-1, (b-1)(k-1)}$
- p-value: $P(F_{k-1, (b-1)(k-1)} \geq F_{obs})$

 
The procedures to make comparisons among means are
very similar to the methods used for the CRD. In each formula
described previously for 
Bonferroni's, and Tukey's methods, $n_i$ is replaced 
by $b$, when making comparisons among treatment means, and  $\nu=(b-1)(k-1)$ is the error degrees of freedom.



**Example 7.7: Comparison of 3 Methods for Estimating Value of Wood Logs**

A study compared 3 methods of assessing the lumber value of logs [@lin_2012]. The $k=3$ treatments the authors compared
was the actual sawmill value of the log, a value based on a heuristic programming algorithm, and a value based on a dynamic programming algorithm.
The "treatments" were each measured on $b=30$ logs (acting as the blocks). The goal was to compare the 3 treatments at valuating the logs. 
Data are given in Table \@ref(tab:logvalue1). A crude interaction plot is given in Figure \@ref(fig:logvalue2), which plots
the valuation versus log ID, with separate lines for the three methods.

```{r logvalue1, echo=FALSE, tidy=FALSE}

lv <- read.csv("sawmill1.csv")
head(lv, 2)

## Structure the data for tabular form where columns represent trts
##  and rows represent blocks

lv1.out <- cbind(lv$lumBlk[1:30], lv$lumVal[1:30], lv$lumVal[31:60], lv$lumVal[61:90])
#head(lv1.out); tail(lv1.out)

Block.mean <- rowMeans(lv1.out[,2:4])
#head(Block.mean)
Trt.mean <- c(NA,colMeans(lv1.out[,2:4]),mean(Block.mean))
Trt.mean 

lv1.out <- rbind(cbind(lv1.out, Block.mean), Trt.mean)
#head(lv1.out); tail(lv1.out)
colnames(lv1.out) <- c("Log ID", "Trt1", "Trt2", "Trt3", "Log Mean")
#head(lv1.out); tail(lv1.out)

knitr::kable(lv1.out, digits=3,
             caption='Measurements, Block Means, and Treatment Means for log valuation experiment',
             booktabs=TRUE)

```
 
```{r logvalue2, fig.cap='Log Valuation vs Log ID with separate lines by treatment (valuation method)', out.width='80%', fig.asp=.75, fig.align='center', echo=FALSE}

## Continuation of logvalue1

interaction.plot(lv$lumBlk, lv$lumTrt, lv$lumVal, main="Log Valuation versus ID by Method",
   pch=16, col=c("purple", "green", "red"), lty=1, lwd=2,
   xlab="Log ID", ylab="Valuation")

```




The calculations for the Analysis of Variance are given below.

 
$$ TSS=(17.67-26.3730)^2 + \cdots + (26.54-26.3730)^2 = 5170.073$$    $$ df=30(3)-1 = 89  $$
$$ SST = 30\left[(24.8743-26.3730)^2+(26.8120-26.3730)^2+(27.4327-26.3730)^2\right] = 106.8536 $$
$$df_T=3-1=2 \qquad MST=\frac{106.8536}{2} = 53.4268$$
$$ SSB = 3\left[(19.8433-26.3730)^2 + \cdots + (25.7700-26.3730)^2\right] = 5042.772$$ 
$$  df_B=30-1=29 \qquad MSB=\frac{5042.772}{29}=173.89 $$
$$ SSE = (17.67-19.8433-24.8743+26.3730)^2+ \cdots +(26.54-25.7700-26.4327+26.3730)^2 = $$
$$ =5170.073-106.8536-5042.772=20.448 = SSE $$
$$ df_E=(30-1)(3-1)=58\qquad MSE=\frac{20.448}{58}=0.353 $$


 
We can now test for treatment effects, and if necessary use Tukey's method
to make pairwise comparisons among the three methods ($\alpha_E=0.05$ significance
level). 

- $H_0: \alpha_1 = \alpha_2 = \alpha_3 = 0  \quad (\mu_1 = \mu_2 = \mu_3)$
 (No differences among valuation method means)
- $H_A: \mbox{ Not all } \alpha_i=0$  (Valuation differences exist)
- T.S. $F_{obs} = \frac{MST}{MSE}=151.546$
- R.R.: $F_{obs} \geq F_{\alpha, k-1, (b-1)(k-1)}=F_{0.05,2,58}=3.156$
- p-value: $P=P(F_{2,58} \geq F_{obs})=P(F_{2,58}\geq 151.546)=0.0000$

 
Tukey's method is used to determine which valuation methods
differ significantly. Recall that for Tukey's method, 
simultaneous confidence intervals of the form given below are computed, with
$k$ being the number of treatments ($k$=3), $b$ being the number of blocks, and $n_i$ the number of measurements per
valuation method ($n_i=b=30$).

$$ \left(\overline{y}_{i.} - \overline{y}_{i'.}\right) \pm q_{\alpha,k, (b-1)(k-1)} \sqrt{MSE\left(\frac{1}{n_i}\right)} $$
$$\quad \Longrightarrow \quad 
 \left(\overline{y}_{i.} - \overline{y}_{i'.}\right) \pm 3.402 \sqrt{0.353\left(\frac{1}{30}\right)} \quad
\Longrightarrow \left(\overline{y}_{i.} - \overline{y}_{i'.}\right) \pm 0.369 
$$ 

We fit the Analysis of Variance using the **aov** function in R and 
obtain the Tukey simultaneous Confidence Intervals in the following R code chunk.

```{r logvalue3, echo=FALSE}

## Continuation of logvalue1
lv$lumTrt.f <- factor(lv$lumTrt, levels=1:3,
                      labels=c("Actual", "Heuristic", "Dynamic"))
lv$lumBlk.f <- factor(lv$lumBlk)

lv.mod1 <- aov(lumVal ~ lumTrt.f + lumBlk.f, data=lv)
anova(lv.mod1)
TukeyHSD(lv.mod1, "lumTrt.f")

```


Conclude that Actual sawmill valuation is significantly lower than Heuristic, which is significantly lower than Dynamic.

Note that to run this in R, it is necessary to have a separate row for each observation, along with a treatment ID and block ID (as was used in the previous R code chunk).

$$\nabla $$


## Factorial Designs

In many cases, the research is interested in whether multiple factors have effects on responses, and whether
the effects of the individual factor levels depend on the levels of the remaining factor(s). We will consider only the case of two
treatment factors: $A$ with $a$ levels and $B$ with $b$ levels. Further, we will only consider the case where there are $n$
observation within each of the $ab$ treatments (the crossing of the levels of factors $A$ and $B$).

In this section, models with two factors are considered.  
Denoting the $k^{th}$ measurement
observed under the $i^{th}$ level of factor $A$ and the $j^{th}$ level
of factor $B$, the  
model is written as follows.

$$ Y_{ijk} = \mu_{ij} + \epsilon_{ijk} = \mu + \alpha_i + \beta_j + (\alpha \beta)_{ij} + \epsilon_{ijk}  $$ $$i=1,\ldots,a;\quad j=1,\ldots,b;\quad  k=1,\ldots,n \qquad \epsilon_{ijk}\sim 
N\left(0,\sigma\right)  $$

Here $\mu$ is the overall mean, $\alpha_i$ is the effect of the
$i^{th}$ level of factor $A$, $\beta_j$ is the effect of the $j^{th}$
level of factor $B$, $(\alpha \beta)_{ij}$ is the effect of the 
interaction of the $i^{th}$
level of factor $A$ and the $j^{th}$ level of factor $B$, and
$\epsilon_{ijk}$ is the random error term representing the fact that 
units within each treatment combination will vary, as well as
if the same unit were measured repeatedly, its measurements would vary. 
Here, we consider the model where both factors $A$ and $B$ are **fixed**,
with all levels of interest present in the study.

Note that an interaction represents the fact that the effect of a particular level of factor $A$ depends on the level of factor $B$, and vice versa.
As before, we assume
that $\epsilon_{ijk}$ is normally distributed with mean 0 and variance
$\sigma^2$. 

When factors $A$ and $B$ are fixed, the effects are unknown parameters to be estimated. One common way of parameterizing the model is as follows.

$$ E\left\{Y_{ijk}\right\}=\mu + \alpha_i + \beta_j + (\alpha \beta)_{ij} \qquad\qquad V\left\{Y_{ijk}\right\}=\sigma^2 $$
$$ \sum_{i=1}^a\alpha_i=\sum_{j=1}^b\beta_j=\sum_{i=1}^a(\alpha \beta)_{ij}=\sum_{j=1}^b(\alpha \beta)_{ij}=0 $$


Some interesting hypotheses to test are as follow.

- $H_0^{AB}: (\alpha \beta)_{11} = \cdots = (\alpha \beta)_{ab} = 0$ (No
interaction effect).
- $H_0^A: \alpha_1 = \cdots = \alpha_a =0$ (No effects among the
levels of factor $A$) 
- $H_0^B: \beta_1 = \cdots = \beta_b =0$ (No effects among the
levels of factor $B$) 

 
The total variation in the set of observed measurements can be
decomposed into four parts: variation in the means of the levels of
factor $A$, variation in the means of the levels of factor $B$,
variation due to the interaction of factors $A$ and $B$, and error
variation. The formulas for the sums of squares are 
given here (but will not be pursued further in this course).

$$\overline{y}_{ij.} = \frac{\sum_{k=1}^{n}y_{ijk}}{n} \qquad \qquad
\overline{y}_{i..} = \frac{\sum_{j=1}^{b}\sum_{k=1}^{n}y_{ijk}}{bn} $$
$$ \overline{y}_{.j.} = \frac{\sum_{i=1}^{a}\sum_{k=1}^{n}y_{ijk}}{an} \qquad \qquad
\overline{y}_{...} = \frac{\sum_{i=1}^{a}\sum_{j=1}^{b}\sum_{k=1}^{n}y_{ijk}}{abn} $$
$$ N = abn \qquad \qquad
s_{ij}^2 = \frac{\sum_{k=1}^n\left(y_{ijk}-\overline{y}_{ij.}\right)^2}{n-1} $$

$$\mbox{Total SS: } \quad TSS =\sum_{i=1}^a \sum_{j=1}^{b}\sum_{k=1}^n  \left(y_{ijk} - \overline{y}_{...} \right)^2 \qquad df_{\mbox{Tot}}=N-1 $$

$$ \mbox{Factor A SS: } \quad SSA =
bn\sum_{i=1}^a  \left(\overline{y}_{i..} - \overline{y}_{...} \right)^2 
\quad df_A=a-1   \quad MSA=\frac{SSA}{a-1}$$

$$ \mbox{Factor B SS: } \quad SSB =
an\sum_{j=1}^b  \left(\overline{y}_{.j.} - \overline{y}_{...} \right)^2 
\quad df_B=b-1 \quad MSB=\frac{SSB}{b-1} $$

$$ \mbox{Interaction AB SS: } \quad SSAB =
n\sum_{i=1}^a \sum_{j=1}^{b}  \left(\overline{y}_{ij.}-\overline{y}_{i..}-\overline{y}_{.j.}+  \overline{y}_{...} \right)^2 $$
$$ df_{AB}=(a-1)(b-1) \qquad \qquad MSAB = \frac{SSAB}{(a-1)(b-1)}$$

$$ \mbox{Error SS: } \quad SSE = 
\sum_{i=1}^a \sum_{j=1}^{b}\sum_{k=1}^n  \left(y_{ijk} - \overline{y}_{ij.} \right)^2  \quad
 df_E = ab(n-1) \quad MSE = \frac{SSE}{ab(n-1)}$$


The error sum of squares can also be computed from the within cell standard deviations, which is helpful as many research articles
provide the treatment means and standard deviations.

$$ SSE = \sum_{i=1}^a \sum_{j=1}^{b}\sum_{k=1}^n  \left(y_{ijk} - \overline{y}_{ij.} \right)^2 =(n-1) \sum_{i=1}^a \sum_{j=1}^{b}s_{ij}^2 $$


 

Note that this type of analysis, 
is almost always done on a computer. The analysis of variance can be set
up as shown in Table \@ref(tab:facaov1), assuming that $n$ measurements are
made at each combination of levels of the two factors. 

```{r facaov1, echo=FALSE, tidy=FALSE}
aovtable.out <- rbind(
cbind("Factor A", "a-1", "SSA", "MSA=SSA/(a-1)", "FA=MSA/MSE"),
cbind("Factor B", "b-1", "SSB", "MSB=SSB/(b-1)", "FB=MSB/MSE"),
cbind("Interaction AB", "(a-1)(b-1)", "SSAB", "MSAB=SSAB/[(a-1)(b-1)]",
      "FAB=MSAB/MSE"),
cbind("Error", "ab(n-1)", "SSE", "MSE=SSE/[ab(n-1)]", "   "),
cbind("Total", "N-1", "TSS", "   ","   ")
)
colnames(aovtable.out) <- c("Source", "df", "SS", "MS", "$F_{obs}$")

knitr::kable(aovtable.out, escape=FALSE, caption='The Analysis of Variance Table for the 2-Factor Fixed Effects Design',
             booktabs=TRUE)

```

 

 
The tests for interactions and for effects of factors $A$ and $B$
involve the three $F$-statistics, and can be conducted as follow. Note that under each of the three null hypotheses,
the corresponding expected mean square in the numerator simplifies to $\sigma^2=E\left\{MSE\right\}$.

- $H_0^{AB}: (\alpha \beta)_{11} = \cdots = (\alpha \beta)_{ab} = 0$ (No
interaction effect).
- $H_A^{AB}:$ Not all $(\alpha \beta)_{ij} = 0$ (Interaction effects
exist)
- T.S.: $F_{AB} = \frac{MSAB}{MSE}$
- R.R.: $F_{AB} \geq F_{\alpha, (a-1)(b-1), ab(n-1)}$
- p-value: $P=P\left(F_{(a-1)(b-1),ab(n-1)} \geq F_{AB}\right)$

Assuming no interaction effects exist, we can test for differences among
the effects of the levels of factor $A$ as follows.

- $H_0^A: \alpha_1 = \cdots = \alpha_a = 0$ (No
factor $A$ effect).
- $H_A^A:$ Not all $\alpha_i = 0$ (Factor $A$ effects
exist)
- T.S. $F_{A} = \frac{MSA}{MSE}$
- R.R.: $F_{A} \geq  F_{\alpha, (a-1), ab(n-1)}$
- p-value: $P=P\left(F_{a-1,ab(n-1)} \geq F_A\right)$

Assuming no interaction effects exist, we can test for differences among
the effects of the levels of factor $B$ as follows.

- $H_0^B: \beta_1 = \cdots = \beta_b = 0$ (No
factor $B$ effect).
- $H_A^B:$ Not all $\beta_j = 0$ (Factor $B$ effects
exist)
- T.S. $F_B = \frac{MSB}{MSE}$
- R.R.: $F_B \geq F_{\alpha, (b-1), ab(n-1)}$
- p-value: $P=P\left(F_{(b-1), ab(n-1)} \geq F_{obs}\right)$


Note that if we conclude interaction effects exist, we usually look at
the individual combinations of factors $A$ and $B$ separately
(as in the Completely Randomized Design), and don't
conduct the last two tests, although software packages print them out. 
 

**Example 7.8: Espresso Foam Index by Temperature and Extraction Pressure**

An experiment was conducted to measure foam index $(Y)$ of espresso samples brewed at $a=3$ levels of temperature (factor $A$: 75C, 85C, 95C)
and $b=2$ extraction pressures (factor $B$: 15bar, 20bar) with $n=9$ replicates per treatment [@masella_2015]. The sample means (standard
deviations), temperature means, pressure means, and overall mean are given in Table \@ref(tab:foamindex1).
Note that at 75C, Foam Index increases by $135.0-113.4=21.6$ when pressure increases from 15 to 20. Similarly, the changes are
11.4 and 26.5 for 85C and 95C, respectively. Those are fairly similar increases, but we will test formally for the interaction below.

```{r foamindex1, echo=FALSE, tidy=FALSE}

fi.out <- rbind(
  cbind("75C", "113.4 (15.0)", "135.0 (27.5)", "124.20"),
  cbind("85C", "102.4 (12.2)", "113.8 (20.5)", "108.10"),
  cbind("95C", "91.1 (12.9)", "117.6 (21.7)", "104.35"),
  cbind("Press Mean", "102.30", "122.13", "112.22")
)

colnames(fi.out) <- c("Temp", "15bar (j=1)", "20bar (j=2)", "Temp Mean")
rownames(fi.out) <- c("(i=1)", "(i=2)", "(1=3)", " ")

knitr::kable(fi.out,
             caption='Espresso foam index mean (sd) by Temperature and Pressure',
             booktabs=TRUE)

```


The sums of squares are set-up below., The Analysis of Variance
is given in Table \@ref(tab:foamindex2).
The Analysis of Variance obtained from the **aov** function is
given in the following R code chunk.

$$ SSA = 2(9)\left[(124.20-112.22)^2+(108.10-112.22)^2+(104.35-112.22)^2\right] = 4004 $$
$$ df_A=3-1=2 \qquad \qquad MSA = \frac{4004}{2}=2002 $$

$$ SSB = 3(9)\left[(102.30-112.22)^2+(122.13-112.22)^2\right] = 5309$$ 
$$ df_B = 2-1 =1 \qquad \qquad MSB=\frac{5309}{1}=5309 $$

$$ SSAB = 9\left[(113.4-124.20-102.30+112.22)^2+\cdots \right] = 534 $$
$$ df_{AB} = 2(1)=2 \qquad MSAB=\frac{534}{2} = 267$$

$$ SSE = (9-1)\left[15.0^2+\cdots +21.7^2\right] = 17501$$
$$df_E=3(2)(9-1) = 48 \qquad \qquad MSE=\frac{17501}{48}=364.60$$

```{r foamindex2, echo=FALSE, tidy=FALSE}

fi2.out <- rbind(
  cbind("Temperature (A)", "2", "4004", "2002", "5.485", ".0072"),
  cbind("Pressure (B)", "1", "5309", "5309", "14.545", ".0004"),
  cbind("Interaction (AB)", "2", "534", "267", "0.732", ".4862"),
  cbind("Error", "48", "17501", "365", "  ", "  "),
  cbind("Total", "53", "27348", "   ", "   ", "   ")
)

colnames(fi2.out) <- c("Source", "$df$", "$SS$", "$MS$", "$F_{obs}$", "$P(>F_{obs})$")

knitr::kable(fi2.out, 
             caption='Analysis of Variance Table for Espresso Foam Index experiment',
             booktabs=TRUE)

etp <- read.csv("espresso2.csv")
head(etp, 2)
etp$tempC.f <- factor(etp$tempC)
etp$prssBar.f <- factor(etp$prssBar)
etp.mod1 <- aov(foamIndx ~ tempC.f*prssBar.f, data=etp)
anova(etp.mod1)
TukeyHSD(etp.mod1, 'tempC.f')
TukeyHSD(etp.mod1, 'prssBar.f')
```


Thus, there are significant main effects for Temperature and Pressure, but the interaction is not significant. The R code chunk and output for 
the analysis  (including Tukey's comparisons among Temperature and Pressure effects individually is given above). Note,
the R program uses more internal decimal places than the computations above. Foam Index is significantly higher at 75C than 85C and 95C.
The Foam Index at 85C and 95C are not significantly different. It is significantly higher at 20bar than 15bar.

An interaction plot is given in Figure \@ref(fig:foamindex3).

```{r foamindex3, fig.cap='Foam Index vs Barometric Pressure with separate lines by Temperature', out.width='80%', fig.asp=.75, fig.align='center', echo=FALSE}

## Continuation of foamindex2

interaction.plot(etp$prssBar, etp$tempC, etp$foamIndx, main="Foam Index  versus Pressure by Temp",
   pch=16, col=c("purple", "green", "red"), lty=1, lwd=2,
   xlab="Pressure", ylab="Foam Index")

```

$$\nabla $$





